---
title: "alt-flash vs flashier: GTEx"
author: "Joonsuk Kang"
date: "2021-12-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# some functions

```{r}
source('code/alt-flash-ebnm-2021-1206.R')
```

# GTEX data

Codes are from Willwerscheid and Stephens (2021, ebnm; https://arxiv.org/pdf/2110.00152.pdf)

```{r}
library(tidyverse)
library(flashier)

url_prefix <- "https://github.com/stephenslab/gtexresults/blob/master/data/" 
gtex_url <- paste0(url_prefix, "MatrixEQTLSumStats.Portable.Z.rds?raw=TRUE") 
colors_url <- paste0(url_prefix, "GTExColors.txt?raw=TRUE")
gtex <- readRDS(gzcon(url(gtex_url)))
strong <- t(gtex$strong.z) # Dataset used by Urbut et al. and Wang & Stephens.
gtex.colors <- read_tsv(colors_url, col_names = c("Tissue", "Hex", "RGB")) %>%
  mutate(Tissue = str_remove_all(Tissue, "[\\(\\)\\-]")) %>%
  mutate(Tissue = str_replace_all(Tissue, " +", "_")) %>%
  pull(Hex, name = Tissue)
gtex.colors <- gtex.colors[rownames(strong)]
```


# subsampling columns

The original matrix has dimensions 44 X 16,069. We sample columns of sample size 100, 500, 1k, 5k and compare flashier and altflash. Given a data matrix, the number of factors $\hat{K}$ estimated from flash is a deterministic number since flash is a deterministic algorithm. We use $\hat{K}+3$ as the initial number of factors in altflash.

```{r}
dim(strong)
set.seed(1)

for (n in c(100, 500, 1000, 5000)){

  colidx <- sample(1:ncol(strong), n, replace=FALSE)
  mat <- strong[,colidx]
  fit.flash <- flash(mat, prior.family=c(as.prior(ebnm::ebnm_point_exponential, sign = 1), 
                                         as.prior(ebnm::ebnm_point_normal)), 
                     backfit=TRUE, verbose.lvl=0)
  fit.alt <- altflash(mat, initial.K=fit.flash$n.factors+3, prior.family=c('point_exponential', 'point_normal'), verbose=0)
  print(paste0("n=",n,"; flash elbo=", round(fit.flash$elbo), "; altflash elbo=", round(last(fit.alt$elbo.vec))))
}
```
