---
title: "binary toy data: hclust (complete linkage)"
author: "Joonsuk Kang"
date: "2021-10-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# summary

* using denoised (soft-thresholding, factors 2-8) matrix, hierarchical clustering with complete linkage (default option for `hclust`) can eaisly find the eight clusters

* if we include the factor 1 (i.e. use soft-thresholding, factors 1-8 matrix), the hierarchical clustering method fails

* the result holds true for sd=0.5 and sd=1


# some functions

```{r}
library(tidyverse)
library(gridExtra)
library(pheatmap)

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}
```


# binary data

```{r}
binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
L0 <- binary_toy$L0
F0 <- binary_toy$F0
```

### when sd=0.5, denoised: using factors 2-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)
plotL(eigY$vectors[,1:10], 'leading 10 eigenvectors') # the leading eigenvector is close to constant...

# denoised matrix Y
dY <- eigY$vectors[,2:8] %*% diag(eigY$values[2:8]-eigY$values[9]) %*% t(eigY$vectors[,2:8])
dY <- dY - min(dY) # make nonnegative

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 2-8")

hc <- hclust(as.dist(-dY))
plot(hc, cex=0.01)
ct <- cutree(hc, 8)
plot(ct)
```

### when sd=0.5, denoised: using factors 1-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)

# denoised matrix Y
dY <- eigY$vectors[,1:8] %*% diag(eigY$values[1:8]-eigY$values[9]) %*% t(eigY$vectors[,1:8])
dY <- dY - min(dY) # make nonnegative

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 1-8")

hc <- hclust(as.dist(-dY))
plot(hc, cex=0.01)
ct <- cutree(hc, 8)
plot(ct)
```


### when sd=1, denoised: using factors 2-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)
plotL(eigY$vectors[,1:10], 'leading 10 eigenvectors') # the leading eigenvector is close to constant...

# denoised matrix Y
dY <- eigY$vectors[,2:8] %*% diag(eigY$values[2:8]-eigY$values[9]) %*% t(eigY$vectors[,2:8])
dY <- dY - min(dY) # make nonnegative

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 2-8")

hc <- hclust(as.dist(-dY))
plot(hc, cex=0.01)
ct <- cutree(hc, 8)
plot(ct)
```

### when sd=1, denoised: using factors 1-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)

# denoised matrix Y
dY <- eigY$vectors[,1:8] %*% diag(eigY$values[1:8]-eigY$values[9]) %*% t(eigY$vectors[,1:8])
dY <- dY - min(dY) # make nonnegative

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 1-8")

hc <- hclust(as.dist(-dY))
plot(hc, cex=0.01)
ct <- cutree(hc, 8)
plot(ct)
```



