---
title: "binary toy: denoising"
author: "Joonsuk Kang"
date: "2021-10-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# summary


+ by soft thresholding eigenvalues and dropping the leading factor, we can successfully denoise XX'
+ our key matrix XX'=(LF'+E)(LF'+E)'=LF'FL'+LF'E'+EFL'+EE'
+ "soft thresholding": the term EE' in XX' is close to $p\sigma^2I_n$, so the diagonal terms of XX' are disproportionately big; we can use soft thresholding eigenvalues to solve the issue
+ "dropping the leading factor": the term LF'E'+EFL' blurs the pattern observed in the true data; dropping the first factor $\lambda_1u_1u_1^T$ ($u_1$ is close to a constant) solves the issue


# some functions


### funciton for plotting three versions of cov matrix

1. original data matrix

2. soft thresholding eigenvalues

3. soft thresholding eigenvalues + drop the first factor

```{r}
library(pheatmap)

pheatmap1 <- function(matY, data){
  pheatmap(matY, cluster_rows = FALSE, cluster_cols = FALSE, 
           main=paste0(data, '; original'))
  }
pheatmap2 <- function(matY, data){
  eigY <- eigen(matY)
  matY <- eigY$vectors[,1:8] %*% diag(eigY$values[1:8]-eigY$values[9]) %*% t(eigY$vectors[,1:8])
  
  pheatmap(matY, cluster_rows = FALSE, cluster_cols = FALSE, 
           main=paste0(data, '; soft thresholding, factors 1-8'))
}
pheatmap3 <- function(matY, data){
  eigY <- eigen(matY)
  matY <- eigY$vectors[,2:8] %*% diag(eigY$values[2:8]-eigY$values[9]) %*% t(eigY$vectors[,2:8])

  pheatmap(matY, cluster_rows = FALSE, cluster_cols = FALSE, 
           main=paste0(data, '; soft thresholding, factors 2-8'))
}
```

### some other functions
```{r}
library(tidyverse)
library(gridExtra)


# plot posterior mean of L and F 
plotLF <- function(A.l, A.f, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             F=c(A.f),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figL
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=F, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figF
  
  grid.arrange(figL, figF, nrow=2, 
               top=title)
}

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}

```




# binary data

```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211020.R")

binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
L0 <- binary_toy$L0
F0 <- binary_toy$F0
```

# when sd=0.5

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
```

### LF'FL'

```{r}
data <- "LF'FL'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) 
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) # identical to the original because the matY is rank 8
pheatmap3(matY, data=data)
```

### LF'FL' + EE'

```{r}
data <- "LF'FL'+EE'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) + E %*% t(E)
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```

### LF'FL' + LF'E' + EFL'

```{r}
data <- "LF'FL'+LF'E' + EFL'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) + L0%*%t(F0)%*%t(E) + E%*%F0%*%t(L0)
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```


### full XX'

```{r}
data <- "XX'"
matY <- (L0%*%t(F0)+E) %*% t((L0%*%t(F0)+E))
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```



# when sd=1

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
```


### LF'FL'

```{r}
data <- "LF'FL'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) 
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) # identical to the original because the matY is rank 8
pheatmap3(matY, data=data)
```

### LF'FL' + EE'

```{r}
data <- "LF'FL'+EE'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) + E %*% t(E)
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```

### LF'FL' + LF'E' + EFL'

```{r}
data <- "LF'FL'+LF'E' + EFL'"
matY <- L0 %*% diag(diag(t(F0)%*%F0)) %*% t(L0) + L0%*%t(F0)%*%t(E) + E%*%F0%*%t(L0)
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```


### full XX'

```{r}
data <- "XX'"
matY <- (L0%*%t(F0)+E) %*% t((L0%*%t(F0)+E))
pheatmap1(matY, data=data)
pheatmap2(matY, data=data) 
pheatmap3(matY, data=data)
```


