---
title: "binary toy data: NCut"
author: "Joonsuk Kang"
date: "2021-10-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# summary

* If our data is generated from a nested structure, for example a tree structure, then Normalized Cut (NCut) would be a good method to apply.

* NCut outperforms the hierarchical clustering (`hclust`) with complete linkage: even with the noisier data (factors 1-8), NCut successfully recovers the pattern, while `hclust` couldn't.

* this is a preliminary implementation of the NCut algorithm. A more careful implementation would adaptively find which chunk to split at each stage and stop when a convergence criterion is satisfied



# some functions

### funstions of NCut

```{r}
compute.Ncut <- function(W, y2, vals, i){
  A <- (y2<=vals[i])
  B <- !A
  cutAB <- sum(W[A,B])
  assocAV <- sum(W[A,])
  assocBV <- sum(W[B,])
  
  Ncut <- cutAB/assocAV + cutAB/assocBV
  return(Ncut)
}

NCut <- function(matY, verbose=0){
  
    rn <- rownames(matY)
  
    W <- matY; diag(W) <- 0
    D <- diag(colSums(W))
    Dinvqsrt <- diag(1/sqrt(colSums(W)))
    eig <- eigen(Dinvqsrt%*%(D-W)%*%Dinvqsrt)
    y2 <- c(Dinvqsrt %*% eig$vectors[,nrow(matY)-1])
    if(verbose==2){plot(y2)}
    
    ### choose a split point
    vals <- sort(unique(y2)); vals <- vals[-length(vals)]
    Ncuts <- rep(0, times=length(vals))
    
    for (i in 1:length(Ncuts)){
      Ncuts[i] <- compute.Ncut(W=W, y2=y2, vals=vals, i=i)
    }

    split.point <- vals[which.min(Ncuts)]
    ggplot()+
      geom_line(aes(x=1:length(Ncuts), y=Ncuts))+
      geom_point(aes(x=which.min(Ncuts), y=Ncuts[which.min(Ncuts)]), col='red')+xlab("")+
      ggtitle(paste0('NCuts by splitting points')) -> temp.fig
    
    if(verbose==2){print(temp.fig)}

    l <- (y2<=split.point)
    
    
    ggplot()+
      geom_line(aes(x=1:length(y2), y=y2, col=factor(l)))+ggtitle(paste0('colored eigvec')) -> temp.fig
    if(verbose==2){print(temp.fig)}
    
    matY[l,!l] <- 0; matY[!l, l] <- 0
    
    if(verbose==1|verbose==2){
      pheatmap(matY, cluster_rows = F, cluster_cols = F, main="after NCut")
    }
    matA <- matY[l,l]
    matB <- matY[!l,!l]
    l <- matrix(as.numeric(l), ncol=1)
    rownames(l) <- rn
    
    out.list <- list(l=l, matA=matA, matB=matB)
    return(out.list)
}

todf <- function(l){data.frame(idx=as.numeric(rownames(l)), L=l)}
aggdf <- function(r.list){
  
  out.df <- todf(r.list[[1]]$l)
  for (i in 2:length(r.list)){
    out.df <- full_join(out.df, todf(r.list[[i]]$l), by='idx')
  }
  out.mat <- as.matrix(out.df)
  out.mat <- out.mat[,-1]
  colnames(out.mat) <- NULL
    return(out.mat)
}
```


### some other functions

```{r}
library(tidyverse)
library(pheatmap)

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}
```


# binary data

```{r}
binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
L0 <- binary_toy$L0
F0 <- binary_toy$F0
```

### when sd=0.5, denoised: using factors 2-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)
plotL(eigY$vectors[,1:10], 'leading 10 eigenvectors') # the leading eigenvector is close to constant...

# denoised matrix Y
dY <- eigY$vectors[,2:8] %*% diag(eigY$values[2:8]-eigY$values[9]) %*% t(eigY$vectors[,2:8])
dY <- dY - min(dY) # make nonnegative (to satisfy as a similarity matrix)
rownames(dY) <- 1:800

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 2-8")

# first cut
r1 <- NCut(dY, verbose=1)

# second cut
r2a <- NCut(r1$matA, verbose=1) # split the first chunk
r2b <- NCut(r1$matB, verbose=1) # split the second chunk

# third cut: cut smaller one of the two 
if(nrow(r2a$matA)> nrow(r2a$matB)){ r3a <- NCut(r2a$matA, verbose=1)
}else{ r3a <- NCut(r2a$matB, verbose=1)} 

if(nrow(r2b$matA)> nrow(r2b$matB)){ r3b <- NCut(r2b$matA, verbose=1)
}else{ r3b <- NCut(r2b$matB, verbose=1)
} 
 
# fourth cut: cut smaller one of the two
if(nrow(r3a$matA)> nrow(r3a$matB)){ r4a <- NCut(r3a$matA, verbose=1)
}else{ r4a <- NCut(r3a$matB, verbose=1)} 

if(nrow(r3b$matA)> nrow(r3b$matB)){ r4b <- NCut(r3b$matA, verbose=1)
}else{ r4b <- NCut(r3b$matB, verbose=1)
} 


# plot result
pheatmap(aggdf(list(r1, r2a, r2b, r3a, r3b, r4a, r4b)), cluster_rows = F, cluster_cols = F,
         main="NCut at each stage: 0 vs 1; NA is colored gray")
```



### when sd=0.5, denoised: using factors 1-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)

# denoised matrix Y
dY <- eigY$vectors[,1:8] %*% diag(eigY$values[1:8]-eigY$values[9]) %*% t(eigY$vectors[,1:8])
dY <- dY - min(dY) # make nonnegative
rownames(dY) <- 1:800

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 1-8")

# first cut
r1 <- NCut(dY, verbose=1)

# second cut
r2a <- NCut(r1$matA, verbose=1) # split the first chunk
r2b <- NCut(r1$matB, verbose=1) # split the second chunk

# third cut: cut smaller one of the two 
if(nrow(r2a$matA)> nrow(r2a$matB)){ r3a <- NCut(r2a$matA, verbose=1)
}else{ r3a <- NCut(r2a$matB, verbose=1)} 

if(nrow(r2b$matA)> nrow(r2b$matB)){ r3b <- NCut(r2b$matA, verbose=1)
}else{ r3b <- NCut(r2b$matB, verbose=1)
} 
 
# fourth cut: cut smaller one of the two
if(nrow(r3a$matA)> nrow(r3a$matB)){ r4a <- NCut(r3a$matA, verbose=1)
}else{ r4a <- NCut(r3a$matB, verbose=1)} 

if(nrow(r3b$matA)> nrow(r3b$matB)){ r4b <- NCut(r3b$matA, verbose=1)
}else{ r4b <- NCut(r3b$matB, verbose=1)
} 

# plot result
pheatmap(aggdf(list(r1, r2a, r2b, r3a, r3b, r4a, r4b)), cluster_rows = F, cluster_cols = F,
         main="NCut at each stage: 0 vs 1; NA is colored gray")
```


### when sd=1, denoised: using factors 2-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)
plotL(eigY$vectors[,1:10], 'leading 10 eigenvectors') # the leading eigenvector is close to constant...

# denoised matrix Y
dY <- eigY$vectors[,2:8] %*% diag(eigY$values[2:8]-eigY$values[9]) %*% t(eigY$vectors[,2:8])
dY <- dY - min(dY) # make nonnegative
rownames(dY) <- 1:800

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 2-8")

# first cut
r1 <- NCut(dY, verbose=1)

# second cut
r2a <- NCut(r1$matA, verbose=1) # split the first chunk
r2b <- NCut(r1$matB, verbose=1) # split the second chunk

# third cut: cut smaller one of the two 
if(nrow(r2a$matA)> nrow(r2a$matB)){ r3a <- NCut(r2a$matA, verbose=1)
}else{ r3a <- NCut(r2a$matB, verbose=1)} 

if(nrow(r2b$matA)> nrow(r2b$matB)){ r3b <- NCut(r2b$matA, verbose=1)
}else{ r3b <- NCut(r2b$matB, verbose=1)
} 
 
# fourth cut: cut smaller one of the two
if(nrow(r3a$matA)> nrow(r3a$matB)){ r4a <- NCut(r3a$matA, verbose=1)
}else{ r4a <- NCut(r3a$matB, verbose=1)} 

if(nrow(r3b$matA)> nrow(r3b$matB)){ r4b <- NCut(r3b$matA, verbose=1)
}else{ r4b <- NCut(r3b$matB, verbose=1)
} 

# plot result
pheatmap(aggdf(list(r1, r2a, r2b, r3a, r3b, r4a, r4b)), cluster_rows = F, cluster_cols = F,
         main="NCut at each stage: 0 vs 1; NA is colored gray")
```

### when sd=1, denoised: using factors 1-8

```{r}
set.seed(1)
E <- matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
X <- L0 %*% t(F0) + E
XXt <- X %*% t(X)

matY <- XXt
eigY <- eigen(matY)

# denoised matrix Y
dY <- eigY$vectors[,1:8] %*% diag(eigY$values[1:8]-eigY$values[9]) %*% t(eigY$vectors[,1:8])
dY <- dY - min(dY) # make nonnegative
rownames(dY) <- 1:800

pheatmap(dY, cluster_rows = F, cluster_cols = F, main="XX', factors 1-8")

# first cut
r1 <- NCut(dY, verbose=1)

# second cut
r2a <- NCut(r1$matA, verbose=1) # split the first chunk
r2b <- NCut(r1$matB, verbose=1) # split the second chunk

# third cut: cut smaller one of the two 
if(nrow(r2a$matA)> nrow(r2a$matB)){ r3a <- NCut(r2a$matA, verbose=1)
}else{ r3a <- NCut(r2a$matB, verbose=1)} 

if(nrow(r2b$matA)> nrow(r2b$matB)){ r3b <- NCut(r2b$matA, verbose=1)
}else{ r3b <- NCut(r2b$matB, verbose=1)
} 
 
# fourth cut: cut smaller one of the two
if(nrow(r3a$matA)> nrow(r3a$matB)){ r4a <- NCut(r3a$matA, verbose=1)
}else{ r4a <- NCut(r3a$matB, verbose=1)} 

if(nrow(r3b$matA)> nrow(r3b$matB)){ r4b <- NCut(r3b$matA, verbose=1)
}else{ r4b <- NCut(r3b$matB, verbose=1)
} 

# plot result
pheatmap(aggdf(list(r1, r2a, r2b, r3a, r3b, r4a, r4b)), cluster_rows = F, cluster_cols = F,
         main="NCut at each stage: 0 vs 1; NA is colored gray")
```
