---
title: "binary toy data: nonneg init + point-exponential prior"
author: "Joonsuk Kang"
date: "2021-10-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# obeservations

* When the true $L$ is a binary matrix, or can be well approximated by a binary matrix, the alt-falsh with the nonnegative initialization + point-exponential prior does not work well. 

* the nonnegative initialization results in estimating too many factors, capturing false signals though with small weights.

* an interesting observation is that the fit with oracle init achieves a lower elbo than the fit with nonnegative initialization; the oracle initialization results in dropping a few factors (losing true signals)

* a binary initialization and/or a binary prior would be able to help finding a good solution



# some functions

### Initialize L (direction) for nonnegative priors

```{r}
# return nonnegative direction estiamtes (L)
INIT.L.NonNeg <- function(Y, verbose=1, max.K=20, tol.scale.ratio=1e-10){

  n <- nrow(Y)

  # residual Y
  Yres <- Y

  # root factor
  u <- eigen(Yres)$vectors[,1]
  u <- u * sign(u[which.max(abs(u))])
  u[u<0] <- 0
  u <- u/sqrt(sum(u^2))
  L <- matrix(u, ncol=1)
  d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
  Yres <- Y - L %*% diag(d, nrow=length(d)) %*% t(L)

  # add factors
  for (k in 2:max.K){
    ## find direction
    u <- eigen(Yres)$vectors[,1]
    u <- u * sign(u[which.max(abs(u))])
    u[u<0] <- 0
    u <- u/sqrt(sum(u^2))

    ## find scale
    L <- cbind(L, matrix(u, ncol=1))
    d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
    if(verbose==1){print(paste0(k,'-th factor fitting; ', 'min/max sacle ratio: ', min(d)/max(d)))}
    if(min(d)/max(d) < tol.scale.ratio) {
      L <- L[,-k]
      d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
      break
    }
    Yres <- Y - L %*% diag(d, nrow=length(d)) %*% t(L)
  }

  if(verbose==1){print(paste0(ncol(L), ' factors fitted'))}
  out.list <- list(L=L)
  return(out.list)
}

```

### some more functions

```{r}
library(tidyverse)
library(gridExtra)

# plot posterior mean of L and F 
plotLF <- function(A.l, A.f, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             F=c(A.f),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figL
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=F, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figF
  
  grid.arrange(figL, figF, nrow=2, 
               top=title)
}

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}


# scale L: quadratic programming approach to find optimal scaling
scaleL.qp <- function(Y, L, scale.min, print.scale=TRUE){

  n <- nrow(L); K <- ncol(L)
  Ls <- matrix(0, nrow=n, ncol=K)
  
  # normalized L: column norm 1
  Ln <- L %*% diag(x=1/sqrt(colSums(L^2)),nrow=K)
  
  D <- crossprod(Ln)^2
  d <- rep(0, K);   for (k in 1:K){d[k] <- as.numeric(t(Ln[,k])%*%Y%*%Ln[,k])}
  qp <- quadprog::solve.QP(Dmat=D, dvec=d, Amat=diag(1,nrow=K), bvec=rep(scale.min,K)) 
  
  scale <- qp$solution
  if(print.scale){print(scale)}

  for (k in 1:K){Ls[,k] <- sqrt(scale[k])*Ln[,k]}
  
  out.list <- list(scale=scale, Ls=Ls)
  
  return(out.list)
}
```


# results

```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211017.R")
```

```{r}
binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
L0 <- binary_toy$L0
F0 <- binary_toy$F0
X0 <- L0 %*% t(F0)
```

```{r}
set.seed(1)
X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X%*%t(X)
Y <- XXt
```

### nonnegative initialization (using eigenvectors of Y)

```{r}
temp.obj <- INIT.L.NonNeg(Y, verbose=1) # initialize directions (L) 
Ls <- scaleL.qp(Y, temp.obj$L, scale.min=1e-8, print.scale=FALSE)
plotL(Ls$Ls, title='nonneg initialized L')

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotL(fit.alt$A.l, 'result: alt-flash with nonneg init')
```

### oracle initialization

```{r}
Ls <- scaleL.qp(Y, L0, scale.min=1e-8)
plotL(Ls$Ls, title='oracle L') 

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotL(fit.alt$A.l, 'result: alt-flash with oracle init')
```


### some more seeds

since the fitted L and F are identical, I report only L

```{r}
for (seed in 2:5){
  
  set.seed(seed)
  X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
  Y <- X %*% t(X)
  
  print(paste0('seed=',seed))
  
  # nonneg init
  temp.obj <- INIT.L.NonNeg(Y, verbose=0)
  Ls <- scaleL.qp(Y, temp.obj$L, scale.min=1e-8, print.scale=FALSE)
  plotL(Ls$Ls, title=paste0('seed=',seed, ': nonneg initialized L'))

  fit.alt <- alt.flash(INIT.L(init.L=Ls$Ls, Y=Y))
  print(fit.alt$scales)
  plotL(fit.alt$A.l, paste0('seed=',seed,': alt-flash w/ nonneg init'))
  
  # oracle init
  fit.alt <- alt.flash(INIT.L(init.L=scaleL.qp(Y, L0, scale.min=1e-8, print.scale=FALSE)$Ls, Y=Y))
  print(fit.alt$scales)
  plotL(fit.alt$A.l, paste0('seed=',seed,': alt-flash w/ oracle init'))
}
```

