---
title: "binary toy data: binary init + binary prior"
author: "Joonsuk Kang"
date: "2021-10-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# observations

## to do


* first factor too strong?
* how to select initial binary L? now, using cutoff 0.5--> need to be more adaptive to the structure



# some functions


### Initialize binary matrix L and scale matrix D

```{r}
# Initialize binary matrix L and scale matrix D
INIT.L.Bin <- function(Y, max.K=20, max.K.search=50, verbose=1, set.scale.min=1e-10){

  n <- nrow(Y)

  # residual Y
  Yres <- Y

  # root factor
  u <- eigen(Yres)$vectors[,1]
  u <- u * sign(u[which.max(abs(u))]) # flip if the one with largest size has a negative sign
  u <- u/max(u) # rescale so that the largest is 1
  u <- as.numeric(u>=0.5) # binarize u with 0.5 as cutoff (u<0 also classified 0)
  L <- matrix(u, ncol=1)
  D <- D.qp(Y, A.l=L, B.l=crossprod(L), 
               A.f=L, B.f=crossprod(L), scale.min=set.scale.min, print.scale=FALSE)
  Yres <- Yres - L %*% D %*% t(L)

  # add factors
  for (k in 2:max.K.search){
    ## find direction
    u <- eigen(Yres)$vectors[,1]
    u <- u * sign(u[which.max(abs(u))]) # flip if the one with largest size has a negative sign
    u <- u/max(u) # rescale so that the largest is 1
    u <- as.numeric(u>=0.5) # binarize u with 0.5 as cutoff (u<0 also classified 0)

    ## find scale
    L <- cbind(L, matrix(u, ncol=1))
    D <- D.qp(Y, A.l=L, B.l=crossprod(L), 
                 A.f=L, B.f=crossprod(L), scale.min=set.scale.min, print.scale=FALSE)
    if(verbose==2){print(paste0(k,'-th factor fitted'))}
    if(min(diag(D))<set.scale.min*10) {
      knull <- (diag(D) < set.scale.min*10)
      L <- L[,!knull]
      D <- D.qp(Y, A.l=L, B.l=crossprod(L),
                   A.f=L, B.f=crossprod(L), scale.min=set.scale.min, print.scale=FALSE)
      if(verbose==2){print(paste0(k,'-th factor dropped'))}
    }
    if(ncol(L) == max.K) break
    Yres <- Y - L %*% D %*% t(L)
  }

  if(verbose==1 | verbose==2){print(paste0(ncol(L), ' factors fitted'))}
  out.list <- list(L=L, D=D)
  return(out.list)
}
```



### some more functions

```{r}
library(tidyverse)
library(gridExtra)

# plot posterior mean of L and F 
plotLF <- function(A.l, A.f, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             F=c(A.f),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figL
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=F, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figF
  
  grid.arrange(figL, figF, nrow=2, 
               top=title)
}

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}


# find optimal scaling D by quadratic programming approach
D.qp <- function(Y, A.l, B.l, A.f, B.f, scale.min, print.scale=TRUE){
  
  n <- nrow(A.l); K <- ncol(A.l)
  
  D <- B.l * B.f
  d <- rep(0, K); for (k in 1:K){d[k] <- as.numeric(t(A.l[,k]) %*% Y %*% A.f[,k]) }
  qp <- quadprog::solve.QP(Dmat=D, dvec=d, Amat=diag(1,nrow=K), bvec=rep(scale.min,K)) 
  
  scale <- qp$solution
  if(print.scale){print(scale)}

  return(diag(scale, nrow=K))  # return the scale matrix D
}
```


# results

```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211020.R")
```

```{r}
binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
L0 <- binary_toy$L0
F0 <- binary_toy$F0
X0 <- L0 %*% t(F0)
```

```{r}
set.seed(1)
X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X%*%t(X)
Y <- XXt
```

### eigendecomposition

```{r}
eigY <- eigen(Y)
plotL(eigY$vectors[,1:12], 'eigvectors of XXt')
eigY$values[1:20]
```


### XXt 

```{r}
# using Y
Y <- XXt
temp.obj <- INIT.L.Bin(Y, max.K=20, max.K.search=50, verbose=1, set.scale.min=1e-10) 
plotL(temp.obj$L, title='initialized binary L')

fit.init <- INIT.L(Y=Y, L=temp.obj$L, D=temp.obj$D, g.pf='binary')
fit.alt <- alt.flash(fit.init, g.common=TRUE)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: binary alt-flash')
```

### XXt, without first factor

```{r}
Y <- eigY$vectors[,-1] %*% diag(eigY$values[-1]) %*% t(eigY$vectors[,-1])
temp.obj <- INIT.L.Bin(Y, max.K=20, max.K.search=50, verbose=1, set.scale.min=1e-10) 
plotL(temp.obj$L, title='initialized binary L')

fit.init <- INIT.L(Y=Y, L=temp.obj$L, D=temp.obj$D, g.pf='binary')
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: binary alt-flash w/o first factor')
```

### XXt, with half-sized F1

```{r}
F0h <- F0
F0h[,1] <- F0h[,1]/2
pheatmap::pheatmap(F0, cluster_rows=FALSE, cluster_cols = FALSE)
pheatmap::pheatmap(F0h, cluster_rows=FALSE, cluster_cols = FALSE)

set.seed(1)
X <- L0 %*% t(F0h) + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X%*%t(X)
Y <- XXt
plotL(eigen(Y)$vectors[,1:12], 'eigvectors of XXt, half-sized F1')

temp.obj <- INIT.L.Bin(Y, max.K=20, max.K.search=50, verbose=1, set.scale.min=1e-10) 
plotL(temp.obj$L, title='initialized binary L')

fit.init <- INIT.L(Y=Y, L=temp.obj$L, D=temp.obj$D, g.pf='binary')
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: binary alt-flash w/o first factor')
```



### XXt, with oracle init

```{r}
# using Y
#Y <- eigY$vectors %*% diag(eigY$values) %*% t(eigY$vectors)
Y <- XXt
D <- D.qp(Y=Y, A.l=L0, B.l=crossprod(L0),
     A.f=L0, B.f=crossprod(L0),
     scale.min=1e-10)
plotL(L0, title='oracle binary L')

fit.init <- INIT.L(Y=Y, L=L0, D=D, g.pf='binary')
fit.alt <- alt.flash(fit.init, g.common=TRUE)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: binary alt-flash with oracle init')
```





### (recap) nonnegative initialization + point-exponential fit

```{r, eval=FALSE}
temp.obj <- INIT.L.NonNeg(Y, verbose=1) # initialize directions (L) 
Ls <- scaleL.qp(Y, temp.obj$L, scale.min=1e-8, print.scale=FALSE)
plotL(Ls$Ls, title='nonneg initialized L')

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y, g.pf='point_exponential')
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotL(fit.alt$A.l, 'result: alt-flash with nonneg init')
```


### some more seeds







### exploring other rules... not completed

```{r, eval=FALSE}
# Initialize binary matrix L and scale matrix D
INIT.L.Bin <- function(Y, max.K=20, max.K.search=50, verbose=1, set.scale.min=1e-10){

  n <- nrow(Y)
  L <- matrix(1, nrow=n, ncol=1)
  D <- D.qp(Yres, A.l=U, B.l=crossprod(U), 
            A.f=U, B.f=crossprod(U), scale.min=set.scale.min, print.scale=FALSE)
  Yres <- Y - L %*% D %*% t(L) # residual Y

  for (k in 2:max.K.search){
    u <- eigen(Yres)$vectors[,1]
    u <- u * sign(u[which.max(abs(u))]) # flip if the one with largest size has a negative sign
    u <- u/max(u) # rescale so that the largest is 1
    u <- as.numeric(u>mean(u[u>0]))
    U <- matrix(u, ncol=1)
    D <- D.qp(Yres, A.l=U, B.l=crossprod(U), 
              A.f=U, B.f=crossprod(U), scale.min=set.scale.min, print.scale=FALSE)
    if(as.numeric(D)<set.scale.min*2) break
    L <- cbind(L, U)
    Yres <- Yres - U %*% D %*% t(U)
  }  
  
  pheatmap::pheatmap(L, cluster_rows=FALSE, cluster_cols=FALSE)

  D <- D.qp(Y, A.l=L, B.l=crossprod(L), 
                 A.f=L, B.f=crossprod(L), scale.min=set.scale.min, print.scale=FALSE)
  keep.idx <- c(1, sort(order(diag(D)[-1], decreasing=TRUE)[1:max.K])+1) # always keep the root factor
  D <- diag(diag(D)[keep.idx])
  L <- L[,keep.idx]
  

  if(verbose==1 | verbose==2){print(paste0(ncol(L), ' factors fitted'))}
  out.list <- list(L=L, D=D)
  return(out.list)
}
```
