---
title: "binary toy data"
author: "Joonsuk Kang"
date: "2021-10-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

I created a binary toy data, available as

```{r,}
binary_toy <- readRDS(url("https://raw.githubusercontent.com/joonsukkang/mf/master/data/binary_toy.rds"))
```



# binary toy data

As a benchmark, let's make a binary toy data, meaning $L$ is a binary matrix, from the toy data from Yusha. 

### import the data

```{r}
library(tidyverse)
library(pheatmap)

# source: https://yushaliu.github.io/single_cell_cancer/simulations3_update.html
model <- readRDS(url("https://raw.githubusercontent.com/YushaLiu/single_cell_cancer/master/data/model.rds"))
L <- model$L
F <- model$F
F[,1] <- F[,1] - 10 # center the data

### plot the drift factorization of truth, where F is orthogonal across columns 
pheatmap(L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True L")
pheatmap(F, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")
pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")
```

### transformation

The true $L$ has all nonnegative entries, and the mean of the positive entries is 1. So, it seems fine to transform all the positive entries to 1.

```{r}
summary(c(L))
summary(c(L[L>0]))

L0 <- L
L0[L0>0] <- 1
summary(c(L0))
pheatmap(L0, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Binary L")
F0 <- F
```

### compare the binary version vs original

```{r}
# noiseless data
X0 <- L0 %*% t(F0) # binary version
X  <- L %*% t(F) 

s0 <- svd(X0)
s <- svd(X)
```

The 8 largest singular values (signal part) are almost identical (a little smaller in the binary case).

```{r}
# singular values
s0$d[1:15]
s$d[1:15]
s0$d[1:15]/s$d[1:15]
```

```{r}
# plot singular vectors
plot.L10 <- function(X, title){
  
  s <- svd(X)
  data.frame(L.svd=c(s$u[,1:10]),
             x=rep(1:800,times=10),
             factor=rep(1:10, each=800)) %>%
    mutate(factor=as.factor(factor)) %>%
    ggplot()+
    geom_line(aes(x=x, y=L.svd, group=factor, col=factor))+
    facet_wrap(~factor)+ggtitle(title)
}

plot.L10(X0, "Left singular vector of binary toy data")
plot.L10(X, "Left singular vector of original toy data")
```

### save the data

```{r, eval=FALSE}
out.list <- list(L0=L0, F0=F0)
saveRDS(out.list, "data/binary_toy.rds")
```

