---
title: "The Model with Exponential or Exponential Mixture Prior"
author: "Joonsuk Kang"
date: "2020-11-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
 
# Toy Data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

rm(X, X.svd, f, i, n, p)
```


# Demo Functions

## Initialization

```{r}
demo.init <- function(mat.X, r2, n.grid){
  
  N <- nrow(mat.X)
  P <- ncol(mat.X)
  
  eig.XXt <- eigen(mat.X %*% t(mat.X))
  l <- c(sqrt(r2) %*% (eig.XXt$vectors[,1]))
  w <- l^2
  Ew.inv <- 1/w
  
  K <- n.grid
  if(K==1){ 
            pi.grid <- 1
            eta.grid <- r2/N 
    }
  if(K>1){
            pi.grid <- rep(1/K, times=K)
            eta.grid <- (r2/N)*(K/(2^K-1))*2^seq(0,K-1,by=1)
  }
  
  ElXXl <- t(l) %*% mat.X %*% t(mat.X) %*% l
  sigmasq <- c((norm(mat.X, 'f')^2-ElXXl/r2) * 1/((N-1)*P))
  delta <- c((1/r2)*((ElXXl/r2)/((norm(mat.X, 'f')^2-ElXXl/r2)/N-1)-1))
  
  out.list = list()
  out.list$l <- l
  out.list$w <- w
  out.list$eta.grid <- eta.grid
  out.list$pi.grid <- pi.grid
  out.list$Ew.inv <- Ew.inv
  out.list$sigmasq <- sigmasq
  out.list$delta <- delta
  
  return(out.list)
}
```


## Updates


```{r}
demo.update.A <- function(mat.X, sigmasq, delta, Ew.inv, r2){

    N <- nrow(mat.X)
    P <- ncol(mat.X)
  
    Y <- mat.X %*% t(mat.X)/(sigmasq*(r2+1/delta)) - diag(x=Ew.inv)
    Y.eigen <- eigen(Y)
    
    eval.expectation <- function(k){(sum(1/(k/(sigmasq*(r2+1/delta))-Y.eigen$values))-r2)^2}
    find.k.star <- optimize(f=eval.expectation,
                            interval = c((Y.eigen$values[1]+1/r2)*sigmasq*(r2+1/delta),
                                         (Y.eigen$values[1]+N/r2)*sigmasq*(r2+1/delta))
                            )
    k.star <- find.k.star$minimum
  
    d.tilde <- 1/(k.star/(sigmasq*(r2+1/delta))-Y.eigen$values)
    Sigma.star <- Y.eigen$vectors %*% diag(x=d.tilde) %*% t(Y.eigen$vectors)
    log.det.Sigma.star <- sum(log(d.tilde))

    Elsq <- diag(Sigma.star)
    ElXXl <- sum(diag((t(mat.X) %*% Sigma.star %*% mat.X)))
    
    sigmasq <- c((norm(mat.X, 'f')^2-ElXXl/r2) * 1/((N-1)*P))
    delta <- c((1/r2)*((ElXXl/r2)/((norm(mat.X, 'f')^2-ElXXl/r2)/N-1)-1))

    out.list = list()
    out.list$log.det.Sigma.star = log.det.Sigma.star
    out.list$Sigma.star = Sigma.star
    out.list$Elsq = Elsq
    out.list$ElXXl = ElXXl
    out.list$sigmasq = sigmasq
    out.list$delta = delta
    out.list$l = sqrt(r2)*(Y.eigen$vectors[,1])

    return(out.list)
}
```

```{r}
demo.update.B <- function(mat.X, Elsq, eta.grid, pi.grid, tol.convergence, update.pi.tilde){
  
    N <- nrow(mat.X)
    K <- length(pi.grid)
  
    if(K==1){ eta.grid <- 2*mean(sqrt(Elsq))^2
              Ew.inv <- 1/(sqrt(Elsq)*mean(sqrt(Elsq)))
    }
    
    if(K>1 & (update.pi.tilde=='exp')){ 
      
      # initialize E[w]: replacing pi.tilde with pi
      Ew <- sqrt(1/(2*sum(pi.grid * (1/eta.grid)))) * ( sqrt(Elsq) + sqrt(1/(2*sum(pi.grid * (1/eta.grid)))) ) 
      check.convergence <- 1
      while(check.convergence > tol.convergence){
              
              # update pi.tilde
              pi.tilde <- matrix(0, nrow=N, ncol=K)
              for (k in 1:K){
                pi.tilde[,k] <- pi.grid[k]*dexp(Ew, rate=1/eta.grid[k])
              }    
              for (i in 1:N){ # normalize to sum to 1 for each row  
                pi.tilde[i,] <- pi.tilde[i,]/sum(pi.tilde[i,])   
              }
              
              # update E[w]
              Ew.old <- Ew
              Ew <- c(1/sqrt(2*pi.tilde %*% (1/eta.grid))) * ( sqrt(Elsq) + c(1/sqrt(2*pi.tilde %*% (1/eta.grid))) )
              check.convergence <- sum((Ew-Ew.old)^2)
              #print(check.convergence)
      }
      Ew.inv <- c(sqrt(2/Elsq) * sqrt(pi.tilde %*% (1/eta.grid)))
    }
    
    if(K>1 & (update.pi.tilde=='laplace')){ 
        
        pi.tilde <- matrix(0, nrow=N, ncol=K)
        for (k in 1:K){
          pi.tilde[,k] <- pi.grid[k]/sqrt(2*eta.grid[k])*exp(-sqrt(Elsq*2/pi)/sqrt(eta.grid[k]/2))
        }    
        for (i in 1:N){ # normalize to sum to 1 for each row  
          pi.tilde[i,] <- pi.tilde[i,]/sum(pi.tilde[i,])   
        }
        Ew.inv <- c(sqrt(2/Elsq) * sqrt(pi.tilde %*% (1/eta.grid)))
    }
    
    
    out.list = list()
    out.list$Ew.inv = Ew.inv
    out.list$eta.grid = eta.grid
    if(K>1 ){out.list$pi.tilde = pi.tilde}
    
    return(out.list)
}
```


## initialization + updates

```{r}
demo.fit <- function(mat.X, niter, r2=1, n.grid=15, update.pi.tilde='laplace', tol.convergence=10^(-6)){
  
    N <- nrow(mat.X)
    P <- ncol(mat.X)
  
    par.init <- demo.init(mat.X, r2, n.grid=n.grid) 
                                               
    Ew.inv   <- par.init$Ew.inv
    sigmasq  <- par.init$sigmasq
    delta    <- par.init$delta
    eta.grid <- par.init$eta.grid
    pi.grid  <- par.init$pi.grid

    for (i in 1:niter){
    # leg A
    temp.A <- demo.update.A(mat.X,
                          sigmasq=sigmasq,
                          delta=delta,
                          Ew.inv=Ew.inv,
                          r2=r2)
    Sigma.star         <- temp.A$Sigma.star
    log.det.Sigma.star <- temp.A$log.det.Sigma.star
    Elsq               <- temp.A$Elsq
    ElXXl              <- temp.A$ElXXl
    sigmasq            <- temp.A$sigmasq
    delta              <- temp.A$delta
  
    # leg B
    temp.B <- demo.update.B(mat.X,
                          Elsq=Elsq,
                          eta.grid=eta.grid,
                          pi.grid=pi.grid,
                          update.pi.tilde = update.pi.tilde,
                          tol.convergence = tol.convergence)
    Ew.inv <- temp.B$Ew.inv
    eta.grid <- temp.B$eta.grid
    }
    
    out.list = list()
    out.list$log.det.Sigma.star <- log.det.Sigma.star
    out.list$Sigma.star <- Sigma.star
    out.list$Elsq <- Elsq
    out.list$ElXXl <- ElXXl
    out.list$sigmasq <- sigmasq
    out.list$delta <- delta
    out.list$eta.grid <- temp.B$eta.grid
    out.list$pi.grid <- pi.grid
    if(n.grid>1){out.list$pi.tilde <- temp.B$pi.tilde}
    out.list$Ew.inv <- Ew.inv
    out.list$l <- temp.A$l
    return(out.list)
}
```





# Main Analysis: Exponential Prior {.tabset .tabset-fade}

## `X2`

```{r, fig.height=4, fig.width=12}
mat.X <- X2
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```


## `X2n_a` 

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_a
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```

## `X2n_b`

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_b
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=1) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0,b=0)
```

#


# Main Analysis: Exponential Mixture Prior {.tabset .tabset-fade}

Two sets of results are presented depending on how to update the posterior assignment probability $\tilde{\pi}$. 

First approach (`exp`): 
\[
\tilde{\pi}_{ik}=\frac{\pi_k \exp(E[w_i]; \frac{1}{\eta_k})}{\sum_{k'}\pi_{k'} \exp(E[w_i]; \frac{1}{\eta_{k'})}}
\]

Second approach (`laplace`): 
\[
\tilde{\pi}_{ik}=\frac{\pi_k Laplace(\sqrt{\frac{2}{\pi}E[l_i^2]}; 0,\frac{\eta_k}{2})}{\sum_{k'}\pi_{k'} Laplace(\sqrt{\frac{2}{\pi}E[l_i^2]}; 0,\frac{\eta_{k'}}{2})}
\]




## `X2`+`exp`

```{r, fig.height=4, fig.width=12}
mat.X <- X2
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```


## `X2n_a`+`exp`

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_a
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```

## `X2n_b`+`exp`

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_b
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='exp') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0,b=0)
```


## `X2`+`laplace`

```{r, fig.height=4, fig.width=12}
mat.X <- X2
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```


## `X2n_a`+`laplace`

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_a
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
```

## `X2n_b`+`laplace`

```{r, fig.height=4, fig.width=12}
mat.X <- X2n_b
par(mfrow=c(1,3))

obj.fit    <- demo.fit(mat.X, r2=2, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) # sigma^2, delta, E[l^Tl]

obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15, update.pi.tilde='laplace') 
plot(obj.fit$l, ylab='fitted', main='fitted l')
c(obj.fit$sigmasq, obj.fit$delta, sum(diag(obj.fit$Sigma.star))) 

plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
abline(a=0,b=1); abline(a=0,b=-1); abline(a=0,b=0)
```


#



The second approach (`laplace`) is set as the default in the following analyses.




# The Number of Grid Points {.tabset .tabset-fade}

Currently, default number of grid points is set to **K=15** and each of them has probability 1/K.

## `X2` 

```{r, fig.height=4, fig.width=16}
mat.X <- X2

par(mfrow=c(1,4))
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=5) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=10) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=20) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
```

## `X2n_a`

```{r, fig.height=4, fig.width=16}
mat.X <- X2n_a

par(mfrow=c(1,4))
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=5) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=10) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=20) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
```

## `X2n_b`

```{r, fig.height=4, fig.width=16}
mat.X <- X2n_b

par(mfrow=c(1,4))
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=5) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=10) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=15) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
obj.fit    <- demo.fit(mat.X, r2=1, niter=10, n.grid=20) 
plot(obj.fit$l, ylab='fitted', main='fitted l')
```

#







# More Seeds {.tabset .tabset-fade}

```{r, fig.height=4, fig.width=8}

generate.noiseless <- function(seed.no){
  
    set.seed(seed.no)
  
    # create X
    p = 1000
    n = 20
    f = list()
    for(i in 1:6){ 
      f[[i]] = rnorm(p)
    }
    X =matrix(0,ncol=4*n, nrow=p)
    X[,1:(2*n)] = f[[1]]
    X[,(2*n+1):(4*n)] = f[[2]]
    
    X[,1:n] = X[,1:n]+f[[3]]
    X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
    X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
    X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
    X.svd = svd(X)
    
    # create X2
    X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

    # take transpose
    X2 <- t(X2)

    return(X2)
}
```

## again, `123`

The noiseless data is identical, but the five noisy data are different from previous noisy data.

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=123) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```


## seed `773`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=773) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```


## seed `312`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=312) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```


## seed `82`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=82) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```


## seed `1`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=1) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```

## seed `2`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=2) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```

## seed `3`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=3) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```

## seed `4`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=4) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```

## seed `5`

```{r, fig.height=12, fig.width=16}
X.baseline <- generate.noiseless(seed.no=5) 

par(mfrow=c(3,4))
for (i in 1:6){
  if(i==1){mat.X <- X.baseline} # noiseless baseline
  if(i >1){mat.X <- X.baseline+rnorm(4*20*1000, sd=3)} # noisy versions
  obj.fit <- demo.fit(mat.X, r2=1, niter=10) 

  plot(obj.fit$l, ylab='fitted', main='fitted l')
  plot(x=svd(mat.X)$u[,1], y=obj.fit$l, xlab='singular vector', ylab='fitted', main='fitted l')
  abline(a=0,b=1); abline(a=0,b=-1); abline(a=0, b=0)
}
```

#




