---
title: "Integrate f out: demo2--sprase binary prior"
author: "Joonsuk Kang"
date: "2020-11-3"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

The prior is set as a sparse binary distribution: probability 0.5 at (a value close to zero) and (a large value). When given two points, e.g. $(\kappa_0, \kappa_1)=(10^{-15}, 10^{15})$, the prior distribution for `w` is set as the discrete probability distribution with probability mass 0.5 at each of the two points. The first singular vector of the data is used as the initialization of `l` and The initialization for `w_i` is set using a two-step procedure. First, set $w_i=l_i^2$ using the initialization $l_i$'s. Second, if $w_i$ is less than the median of $\{w_i\}_{i=1}^n$, replace the value with $\kappa_0$; otherwise, replace it with $\kappa_1$. For details, refer to the function `demo.init.svd` below. 

A grid of values $(\kappa_0, \kappa_1)$ are explored. 

This shows that there exist settings in which the method works.

Minor fixes:

+ elbo calculation
+ use evd in update steps instead of svd


## Data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

# svd
X2.svd = svd(X2)
X2n_a.svd = svd(X2n_a)
X2n_b.svd = svd(X2n_b)
rm(X)
```

## Demo Functions

### Initialization

```{r}
# initialization of l using svd
demo.init.svd <- function(mat.X, warmstart=FALSE, warmstart.l=NA,
                          grid.bin0, grid.bin1, prob.bin0){
      
    N <- nrow(mat.X)
    P <- ncol(mat.X)
    
    mat.X.svd <- svd(mat.X)
    init.l <- mat.X.svd$u[,1]
    ############### temporary functionality; overwrite init.l if given warmstart
    if(warmstart==TRUE){ init.l <- warmstart.l }
    ###############
    init.w <- init.l^2
    
    init.sigmasq <- c((norm(mat.X, type='f')^2-t(init.l)%*%mat.X%*%t(mat.X)%*%init.l)/((N-1)*P))
    init.delta <- c((t(init.l)%*%mat.X%*%t(mat.X)%*%init.l)/(P*init.sigmasq)-1)
    
    # set the grid
    grid.min <- grid.bin0
    grid.max <- grid.bin1
    sigmasq.grid <- c(grid.min, grid.max)
    
    init.pi <- c(prob.bin0, 1-prob.bin0)
    
    # change init.w: into two groups
    init.w[init.w<quantile(init.w, prob.bin0)] <- grid.min
    init.w[init.w>=quantile(init.w, prob.bin0)] <- grid.max
    
    init.Ew.inv <- 1/init.w
    
    out.list = list()
    out.list$l <- init.l
    out.list$w <- init.w
    out.list$Ew.inv <- init.Ew.inv
    out.list$sigmasq <- init.sigmasq
    out.list$delta <- init.delta
    out.list$pi <- init.pi
    out.list$sigmasq.grid <- sigmasq.grid
    
    return(out.list)
}
```


### Updates


```{r}
# leg A: given w, update l (essentially, Sigma.star), sigmasq, delta
demo.update.A <- function(mat.X, sigmasq, delta, Ew.inv){
    # input: mat.X; sigmasq, delta, Ew.inv, 
    # output: Sigma.star, Elsq, Elxxl; updated sigmasq, delta; u1 (first left singular vector of Y)
  
    N <- nrow(mat.X)
    P <- ncol(mat.X)
  
    Y <- mat.X%*%t(mat.X)/(sigmasq*(1+1/delta)) - diag(x=Ew.inv)
    Y.eigen <- eigen(Y)
    
    eval.expectation <- function(k){(sum(1/(k/(sigmasq*(1+1/delta))-Y.eigen$values))-1)^2}
    find.k.star <- optimize(f=eval.expectation,
                            interval = c((Y.eigen$values[1]+1)*sigmasq*(1+1/delta), (Y.eigen$values[1]+N)*sigmasq*(1+1/delta))
                            )
    k.star <- find.k.star$minimum
  
    d.tilde <- 1/(k.star/(sigmasq*(1+1/delta))-Y.eigen$values)
    Sigma.star <- Y.eigen$vectors %*% diag(x=d.tilde) %*% t(Y.eigen$vectors)
    log.det.Sigma.star <- sum(log(d.tilde))

    Elsq <- diag(Sigma.star)
    ElXXl <- sum(diag((t(mat.X) %*% Sigma.star %*% mat.X)))
    
    sigmasq <- (norm(mat.X, type='f')^2-ElXXl/(1+1/delta))/(N*P)
    delta <- max(ElXXl/(P*sigmasq)-1,10^(-10)) # lower bound at 10^(-10)
    
    out.list = list()
    out.list$log.det.Sigma.star = log.det.Sigma.star
    out.list$Sigma.star = Sigma.star
    out.list$Elsq = Elsq
    out.list$ElXXl = ElXXl
    out.list$sigmasq = sigmasq
    out.list$delta = delta
    out.list$u1 = Y.eigen$vectors[,1]

    
    return(out.list)
}


# leg B: update w (essentially, pi_q) and g (essentially, Ew.inv)
demo.update.B <- function(mat.X, Elsq, pi, sigmasq.grid, niter.wg = 10){
    # input: Elsq, pi; sigmasq.grid, niter.wg
    # output: pi, pi_q Ew.inv
  
    N <- nrow(mat.X)
    P <- ncol(mat.X)
  
    for (wg.iter in 1:niter.wg){ # more than one iteration helps
    pi_q <- matrix(data=0, nrow=N, ncol=length(sigmasq.grid))
    for (i in 1:N){
      pi_q[i,] <- pi*sigmasq.grid^(-0.5)*exp(-Elsq[i]/(2*sigmasq.grid))
      pi_q[i,] <- pi_q[i,]/sum(pi_q[i,])
    }
    pi <- colSums(pi_q)/N
    }
    
    Ew.inv <- vector(mode='numeric', length=N)
    for (i in 1:N){
    Ew.inv[i] <- sum(pi_q[i,]*(1/sigmasq.grid))
    }
    
    out.list = list()
    out.list$pi = pi
    out.list$pi_q = pi_q
    out.list$Ew.inv = Ew.inv
    
    return(out.list)
}

demo.elbo <- function(mat.X, N,P,
                      sigmasq, delta,
                      log.det.Sigma.star, ElXXl, Elsq,
                      pi, pi_q, 
                      sigmasq.grid,
                      Ew.inv
                      ){
  logpi <- log(pi)
  logpi[pi==0] <- 0 # to exclude -Inf (where pi=0) in the sum

  elbo <- -N*P*log(sigmasq)/2-P*log(1+delta)/2-norm(mat.X,'f')^2/(2*sigmasq)+ElXXl/(2*sigmasq*(1+1/delta))+
    log.det.Sigma.star/2-sum(log(pi_q)*pi_q, na.rm=TRUE)-sum(pi_q %*% log(sigmasq.grid))/2-sum(Elsq*Ew.inv)/2+
    sum(pi_q %*% logpi)
  
  return(elbo)
}
```


### initialization + updates + elbo

```{r}
demo.fit <- function(mat.X, niter, print.elbo=TRUE, warmstart.l=NA,
                     binary_small=10^(-15), binary_big=10^15, prob_small=0.5 # (prob_big=1-prob_small)
                     ){
  
    N <- nrow(mat.X)
    P <- ncol(mat.X)
  
    par.init <- demo.init.svd(mat.X,
                              grid.bin0=binary_small,    grid.bin1=binary_big,
                                               prob.bin0=prob_small) 
                                               
    l <- par.init$l
    w <- par.init$w
    Ew.inv <- par.init$Ew.inv
    sigmasq <- par.init$sigmasq
    delta <- par.init$delta
    pi <- par.init$pi
    sigmasq.grid <- par.init$sigmasq.grid
    
    elbo <- vector('numeric',length=niter)
    
    for (i in 1:niter){
    # leg A
    temp.A <- demo.update.A(mat.X,
                          sigmasq=sigmasq,
                          delta=delta,
                          Ew.inv=Ew.inv)
    Sigma.star <- temp.A$Sigma.star
    log.det.Sigma.star <- temp.A$log.det.Sigma.star
    Elsq <- temp.A$Elsq
    ElXXl <- temp.A$ElXXl
    sigmasq <- temp.A$sigmasq
    delta <- temp.A$delta
  
    # leg B
    temp.B <- demo.update.B(mat.X,
                          Elsq=Elsq,
                          pi=pi,
                          sigmasq.grid=sigmasq.grid)
    pi <- temp.B$pi
    pi_q <- temp.B$pi_q
    Ew.inv <- temp.B$Ew.inv
    
    # elbo
    elbo[i] <- demo.elbo(mat.X,N,P,
                      sigmasq, delta,
                      log.det.Sigma.star, ElXXl, Elsq,
                      pi, pi_q, 
                      sigmasq.grid,
                      Ew.inv
                      )
    if(print.elbo==TRUE){print(elbo[i])}
    }
    
    out.list = list()
    out.list$log.det.Sigma.star <- log.det.Sigma.star
    out.list$Sigma.star <- Sigma.star
    out.list$Elsq <- Elsq
    out.list$ElXXl <- ElXXl
    out.list$sigmasq <- sigmasq
    out.list$delta <- delta
    out.list$pi <- pi
    out.list$pi_q <- pi_q
    out.list$Ew.inv <- Ew.inv
    out.list$u1 <- temp.A$u1
    out.list$sigmasq.grid <- sigmasq.grid
    out.list$elbo <- elbo
    return(out.list)
}
```




## Demo Analysis

```{r}
library(tidyverse)
library(flashr)
```

### Recap: the flash/svd fit for the rank-1 problem

```{r}
# flash fit: restricted to rank-1
X2.flash1 = flashr::flash(X2,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_a.flash1 = flashr::flash(X2n_a,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_b.flash1 = flashr::flash(X2n_b,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)

data.frame(idx = 1:80,
           l = c(X2.svd$u[,1],
                 X2n_a.svd$u[,1],
                 X2n_b.svd$u[,1],
                 X2.flash1$ldf$l[,1],
                 X2n_a.flash1$ldf$l[,1],
                 X2n_b.flash1$ldf$l[,1]),
           data = rep(rep(c("X2", "X2n_a (hard)", "X2n_b (easy)"), each=80), times=2),
           method = rep(c("svd", "flash"), each=80*3)
           ) %>% 
  # adjust sign for illustration
  mutate(l = ifelse((data=="X2n_b (easy)" & method=="flash"),
                            -l,
                            l)
         ) -> X2.fit1

# warmstart: sparse fit (from X2n_b) initialization for flash on data X2
X2.flash.warmstart1 = flashr::flash(X2,K=1,f_init=X2n_b.flash1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant",backfit = TRUE,greedy = FALSE, verbose=FALSE)
data.frame(idx=1:80, 
           l = c(X2.flash.warmstart1$ldf$l[,1]), data="X2", method="flash_warm") %>%
  rbind(X2.fit1) -> X2.fit1

X2.fit1 %>%
  ggplot()+geom_line(aes(x=idx, y=l, group=method, col=method))+
  facet_grid(method~data, labeller=label_both)+geom_hline(yintercept=0)+
  ggtitle("The Rank-1 Problem")
```

### Analysis for the noiseless data `X2`

```{r, fig.height = 8, fig.width = 10}
# sparse binary prior: =bin0 wp 0.5 and =bin1 wp 0.5
summary((X2.svd$u[,1])^2)

# bin0: essentially 0
# bin1: a large value
niter <- 5
bin0.grid <- 10^c(-15,-10,-7,-5,-4,-3)
bin1.grid <- 10^c(-2,-1,0,1,2,5,10,15)

df.result <- data.frame()
df.elbo <-data.frame()

for (i in 1:length(bin0.grid)){
  for (j in 1:length(bin1.grid)){

          bin0 <- bin0.grid[i]
          bin1 <- bin1.grid[j]

           X2.fit    <- demo.fit(X2, niter=niter, print.elbo=FALSE, binary_small=bin0, binary_big=bin1)
           df.result <- rbind(df.result,
                          data.frame(l=X2.fit$u1,
                                     idx=1:80, bin0 = bin0, bin1=bin1, data="X2"))
           df.elbo <- rbind(df.elbo,
                            data.frame(elbo=c(last(X2.fit$elbo)),
                                       data=c("X2"),
                                       bin0=bin0, bin1=bin1))
  }
}

df.result %>% mutate(bin0=factor(bin0), bin1=factor(bin1)) -> df.result
df.elbo %>% mutate(bin0=factor(bin0)) -> df.elbo

df.result %>% filter(data=="X2") %>%
  ggplot()+geom_line(aes(x=idx, y=l))+geom_hline(yintercept=0, col='red', alpha=0.5)+
  facet_grid(bin0~bin1, labeller='label_both', scales='free')+ggtitle("svd initialization + X2")
```
```{r}
df.elbo %>% filter(data=="X2") %>%
  ggplot(aes(x=bin1, y=elbo, group=bin0, col=bin0))+geom_line()+geom_point(position="jitter")+scale_x_log10()
```


### Analysis for the noiseless data `X2n_a` (hard) and `X2n_b` (easy)


```{r, fig.height = 8, fig.width = 10}
# sparse binary prior: =bin0 wp 0.5 and =bin1 wp 0.5
summary((X2n_a.svd$u[,1])^2)
summary((X2n_b.svd$u[,1])^2)

# bin0: essentially 0
# bin1: a large value
niter <- 5
bin0.grid <- 10^c(-15,-10,-7,-5,-4,-3)
bin1.grid <- 10^c(-2,-1,0,1,2,5,10,15)

df.result <- data.frame()
df.elbo <-data.frame()

for (i in 1:length(bin0.grid)){
  for (j in 1:length(bin1.grid)){

          bin0 <- bin0.grid[i]
          bin1 <- bin1.grid[j]

           X2n_a.fit    <- demo.fit(X2n_a, niter=niter, print.elbo=FALSE, binary_small=bin0, binary_big=bin1)
           X2n_b.fit    <- demo.fit(X2n_b, niter=niter, print.elbo=FALSE, binary_small=bin0, binary_big=bin1)
           df.result <- rbind(df.result,
                          data.frame(l=X2n_a.fit$u1,
                                     idx=1:80, bin0 = bin0, bin1=bin1, data="X2n_a (hard)"),
                          data.frame(l=X2n_b.fit$u1,
                                     idx=1:80, bin0 = bin0, bin1=bin1, data="X2n_b (easy)"))
           df.elbo <- rbind(df.elbo,
                            data.frame(elbo=c(last(X2n_a.fit$elbo), last(X2n_b.fit$elbo)),
                                       data=c("X2n_a (hard)", "X2n_b (easy)"),
                                       bin0=bin0, bin1=bin1))
  }
}

df.result %>% mutate(bin0=factor(bin0), bin1=factor(bin1)) -> df.result
df.elbo %>% mutate(bin0=factor(bin0)) -> df.elbo

df.result %>% filter(data=="X2n_a (hard)") %>%
  ggplot()+geom_line(aes(x=idx, y=l))+geom_hline(yintercept=0, col='red', alpha=0.5)+
  facet_grid(bin0~bin1, labeller='label_both', scales='free')+ggtitle("svd initialization + X2n_a (hard)")
```
```{r}
df.elbo %>% filter(data=="X2n_a (hard)") %>%
  ggplot(aes(x=bin1, y=elbo, group=bin0, col=bin0))+geom_line()+geom_point(position="jitter")+scale_x_log10()
```
```{r, fig.height = 8, fig.width = 10}
df.result %>% filter(data=="X2n_b (easy)") %>%
  ggplot()+geom_line(aes(x=idx, y=l))+geom_hline(yintercept=0, col='red', alpha=0.5)+
  facet_grid(bin0~bin1, labeller='label_both', scales='free')+ggtitle("svd initialization + X2n_b (easy)")
```
```{r}
df.elbo %>% filter(data=="X2n_b (easy)") %>%
  ggplot(aes(x=bin1, y=elbo, group=bin0, col=bin0))+geom_line()+geom_point(position="jitter")+scale_x_log10()
```
