---
title: "Hierarchical ebmf"
author: "Joonsuk Kang"
date: "Feb 4, 2021"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# toy data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000 # original toy data setting: p=1000, n=20
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

rm(X, X.svd, f, i, n, p)
```


# Building Blocks

## `EBNV` solver

```{r}
EBNV <- function(b, G, eta.grid=NULL){
  
  p <- length(b)
  K <- length(eta.grid)
  
  if(G=="exp"){ # exponential family
    
    eta <- 2*mean(abs(b))^2
    EWinv <- sqrt(2/eta)*diag(x=1/abs(b))
    EW <- diag(sqrt(eta/2)*abs(b) + eta/2)

    out.list <- list(EWinv=EWinv, EW=EW, 
                     eta=eta)
  }
  
  if(G=="point.mix"| G=="exp.mix"){  # exponential scale mixture or point mass scale mixture

    L <- matrix(0, nrow=p, ncol=K)
    for (k in 1:K){
      eta.k <- eta.grid[k]
      if(G=="point.mix"){L[,k] <- sqrt(1/eta.k)*exp(-b^2/(2*eta.k))}
      if(G=="exp.mix"  ){L[,k] <- sqrt(pi/eta.k)*exp(-sqrt(2/eta.k)*abs(b))}
    }
    pi.star <- mixsqp::mixsqp(L, control=list(verbose=FALSE))$x
    pi.tilde <- matrix(0, nrow=p, ncol=K)
    for (k in 1:K){
      pi.tilde[,k] <- pi.star[k]*L[,k]
    }
    pi.tilde <- pi.tilde/rowSums(pi.tilde)
    
    if(G=="point.mix"){
      EWinv <- diag(as.numeric(pi.tilde %*%(1/eta.grid)))
      EW <- diag(as.numeric(pi.tilde %*%(eta.grid)))
    }
    if(G=="exp.mix"){
      EWinv <- diag(as.numeric(pi.tilde %*% sqrt(2/eta.grid))/abs(b))
      EW <- diag(as.numeric(pi.tilde %*% sqrt(eta.grid/2))*abs(b) + as.numeric(pi.tilde %*% (eta.grid/2)))
    }

    out.list <- list(EWinv=EWinv, EW=EW, L=L,
                     pi.star=pi.star, pi.tilde=pi.tilde)

  }
  
  if(G=="exp.mix+point.mix"){  # exponential scale mixture + point mass scale mixture

    L <- matrix(0, nrow=p, ncol=2*K)
    for (k in 1:K){
      eta.k <- eta.grid[k]
      L[,k] <- sqrt(1/eta.k)*exp(-b^2/(2*eta.k)) # for point mixture
      L[,(K+k)] <- sqrt(pi/eta.k)*exp(-sqrt(2/eta.k)*abs(b)) # for exp mixture
    }
    pi.star <- mixsqp::mixsqp(L, control=list(verbose=FALSE))$x
    pi.tilde <- matrix(0, nrow=p, ncol=2*K)
    for (k in 1:(2*K)){
      pi.tilde[,k] <- pi.star[k]*L[,k]
    }
    pi.tilde <- pi.tilde/rowSums(pi.tilde)
    
    
    EWinv <- diag(
      as.numeric(pi.tilde[,1:K] %*%(1/eta.grid)) +
        as.numeric(pi.tilde[,(K+1):(2*K)] %*% sqrt(2/eta.grid))/abs(b)
      )

    EW <- diag(
       as.numeric(pi.tilde[,1:K] %*%(eta.grid)) +
         as.numeric(pi.tilde[,(K+1):(2*K)]%*% sqrt(eta.grid/2))*abs(b) +
         as.numeric(pi.tilde[,(K+1):(2*K)] %*% (eta.grid/2))
       )
    

    out.list <- list(EWinv=EWinv, EW=EW, L=L,
                     pi.star=pi.star, pi.tilde=pi.tilde)

  }  
  
  return(out.list)
}
```


## `EBMF-Bingham` solver

### Intermediate functions for calculating two parts ("G1" and "G2")

```{r}
calc.G1 <- function(k, n, p, r2, Xfsq){ # calculate "G1" part of the equation
    sigmasq <- ((n*p*k + Xfsq)-
                       sqrt((n*p*k + Xfsq)^2-4*p*(n-1)*k*Xfsq)
                     )/(2*p*(n-1))
    delta <- sigmasq/(k-sigmasq)/r2
    G1 <- -n*p/2*log(sigmasq)-
          p/2*log(1+delta*r2)-
          Xfsq/(2*sigmasq)
return(c(sigmasq, delta, G1))
}

calc.G2 <- function(k, EWinv, XXT, n, r2){ # calculate "G2" part of the equation
  
  temp.eig <- eigen(EWinv*r2/2-XXT/(2*k))
  eigvals <- temp.eig$values
  lambda1 <- min(eigvals)
  G2 <- -lambda1-1/2*sum(log(1+2*(eigvals-lambda1)/n))+log(2*pi^(n/2))-log(gamma(n/2))
  
  return(G2)
}
```

### Main function

```{r}
EBMF.Bingham <- function(EWinv, XXT, Xfsq, n, p, r2){
  
    sum.G <- function(k){
      calc.G1(k=k, n=n, p=p, r2=r2, Xfsq = Xfsq)[3] + 
        calc.G2(k=k, EWinv=EWinv, XXT=XXT, n, r2=r2)
    }
    
    optimal.k <- optimize(sum.G, lower=0, upper=100, maximum=TRUE, tol=.Machine$double.eps)$maximum
    sigmasq <- ((n*p*optimal.k + Xfsq)-sqrt((n*p*optimal.k + Xfsq)^2-4*p*(n-1)*optimal.k*Xfsq)
                     )/(2*p*(n-1))
    delta <- sigmasq/(optimal.k-sigmasq)/r2
    G2 <- calc.G2(k=optimal.k, EWinv=EWinv, XXT=XXT, n=n, r2=r2)
    
    # computing sqrt.Elsq
    temp.A <- r2/2*EWinv-XXT/(2*sigmasq*(1+1/(delta*r2)))
    tilde.lambda <- eigen(temp.A)$values
    optcondition <- function(c){  (1/2*sum((tilde.lambda+c)^(-1)) -1)^2 }
    optc <- optimize(optcondition, 
                     lower=-min(tilde.lambda)+0.5, 
                     upper= -min(tilde.lambda)+p/2
                     )$minimum
    tilde.mat.approx <- 1/(2*optc)*solve(diag(x=rep(1, times=n))+1/optc*temp.A)
    sqrt.Elsq <- sqrt(r2*diag(tilde.mat.approx))

    out.list <- list(optimal.k=optimal.k, sigmasq=sigmasq, delta=delta, G2=G2, sqrt.Elsq=sqrt.Elsq)

  return(out.list)
}
```


# `hEBMF` solver

```{r}
hEBMF <- function(X, r, G, 
                  min.updates=5, # minimum number of updates
                  fig.intermediate=FALSE, fig.final=TRUE,
                  elbo.tol=10^(-6), rand.init=FALSE){
      
    # setup -----------------------------------------------------------------------
    n <- nrow(X); p <- ncol(X)
    
    Xfsq <- norm(X, 'F')^2
    XXT <- X %*% t(X)
    r2 <- r^2
    XXT.eig <- eigen(XXT)
    XXT.eigvalues <- XXT.eig$values
    l.eigvector <- r*XXT.eig$vectors[,1] # estimate as the eigenvector
    
    # initialization ---------------------------------------------------------------
    sqrt.Elsq <- abs(l.eigvector)
    if(rand.init==TRUE){sqrt.Elsq=sample(sqrt.Elsq)} # permute the eigenvalue to randomize initialization
    
    # create eta grid
    eta.grid <- sort(sqrt.Elsq^2)

    # EBNV
    temp.ebnv <- EBNV(b=sqrt.Elsq, G=G, eta.grid=eta.grid)
    EWinv <- temp.ebnv$EWinv
    EW <- temp.ebnv$EW
    
    # EBMF-Bingham
    temp.ebmf <- EBMF.Bingham(EWinv=EWinv,
                          XXT = XXT, Xfsq=Xfsq, n=n, p=p, r2=r2)
    sigmasq <- temp.ebmf$sigmasq
    delta <- temp.ebmf$delta
    sqrt.Elsq.prev <- sqrt.Elsq
    sqrt.Elsq <- temp.ebmf$sqrt.Elsq
    
    # ELBO 
    if(G!='exp'){
    F1 <- -n*p/(2*pi)-n*p/2*log(sigmasq)-p/2*log(1+delta*r2)-Xfsq/(2*sigmasq)+
      (1/2)*sum(sqrt.Elsq^2*diag(EWinv))+n/2*log(r2)+ temp.ebmf$G2
    F2.F3 <- (-1/2)*sum((sqrt.Elsq^2-sqrt.Elsq.prev^2)*diag(EWinv))+
      sum(log(temp.ebnv$L %*% temp.ebnv$pi.star))-n/2*log(2*pi)
    elbo <- F1+F2.F3
    mat.elbo <- c(F1, F2.F3, elbo)
    }
    
    n.iter <- 1
    par(mfrow=c(1,5))
    # fig.intermediate
    eigen.A <- eigen(-EWinv/2+XXT/(2*sigmasq*(1/delta+r2)))
    l.map <- r*eigen.A$vectors[,1] # point estimate: MAP
    if(fig.intermediate==TRUE){
      plot(l.eigvector, l.map, main=paste0("iteration: ",n.iter))
      abline(a=0,b=0); abline(a=0,b=1); abline(a=0,b=-1)
    }
    
    # updates  ---------------------------------------
    elbo.diff <- 0
    while ( (n.iter < min.updates) | (elbo.diff > elbo.tol) ){
        
        # EBNV
        temp.ebnv <- EBNV(b=sqrt.Elsq, G=G, eta.grid=eta.grid)
        EWinv.prev <- EWinv
        EWinv <- temp.ebnv$EWinv
        EW <- temp.ebnv$EW
        
        # ELBO
        if(G!='exp'){
        F1 <- -n*p/(2*pi)-n*p/2*log(sigmasq)-p/2*log(1+delta*r2)-Xfsq/(2*sigmasq)+
          (1/2)*sum(sqrt.Elsq^2*diag(EWinv.prev))+n/2*log(r2)+ temp.ebmf$G2
        F2.F3 <- sum(log(temp.ebnv$L %*% temp.ebnv$pi.star))-n/2*log(2*pi)
        elbo <- F1+F2.F3
        mat.elbo <- rbind(mat.elbo, c(F1, F2.F3, elbo))    
        }
        
        # EBMF-Bingham
        temp.ebmf <- EBMF.Bingham(EWinv=EWinv, XXT = XXT, Xfsq=Xfsq, n=n, p=p, r2=r2)
        sigmasq <- temp.ebmf$sigmasq
        delta <- temp.ebmf$delta
        sqrt.Elsq.prev <- sqrt.Elsq
        sqrt.Elsq <- temp.ebmf$sqrt.Elsq

        # ELBO
        if(G!='exp'){
        F1 <- -n*p/(2*pi)-n*p/2*log(sigmasq)-p/2*log(1+delta*r2)-Xfsq/(2*sigmasq)+
          (1/2)*sum(sqrt.Elsq^2*diag(EWinv))+n/2*log(r2)+ temp.ebmf$G2
        F2.F3 <- (-1/2)*sum((sqrt.Elsq^2-sqrt.Elsq.prev^2)*diag(EWinv))+
          sum(log(temp.ebnv$L %*% temp.ebnv$pi.star))-n/2*log(2*pi)
        elbo <- F1+F2.F3
        mat.elbo <- rbind(mat.elbo, c(F1, F2.F3, elbo))    
        }
        
        n.iter <- n.iter + 1
        if(G!='exp'){elbo.diff <- mat.elbo[nrow(mat.elbo),3] - mat.elbo[nrow(mat.elbo)-2,3] }
        # compare to last post-EBMF-Bingham elbo
        
        # fig.intermediate
        eigen.A <- eigen(-EWinv/2+XXT/(2*sigmasq*(1/delta+r2)))
        l.map <- r*eigen.A$vectors[,1] # point estimate: MAP
        if(fig.intermediate==TRUE){
          plot(l.eigvector, l.map, main=paste0("iteration: ",n.iter))
          abline(a=0,b=0); abline(a=0,b=1); abline(a=0,b=-1)
        }
     }
     
    # fig.final
    par(mfrow=c(1,4))
    if(fig.final==TRUE){
      if(G!='exp'){plot(mat.elbo[,3], main='ELBO')}
    plot(l.eigvector, main="eigenvector")
    plot(l.map, main="final MAP estimate")
    plot(l.eigvector, l.map, main='eig vs MAP')
    abline(a=0,b=0); abline(a=0,b=1); abline(a=0,b=-1)
    }
    
    # return output
    out.list <- list(l.map = l.map, 
                     sigmasq = sigmasq, 
                     delta = delta, 
                     sqrt.Elsq = sqrt.Elsq,
                     EWinv = EWinv,
                     l.eigvector = l.eigvector, 
                     eigen.A=eigen.A)
    if(G=='exp'){
      out.list <- c(out.list,
                    list(eta=temp.ebnv$eta))
    }
    if(G=='exp.mix' | G=='point.mix' | G=="exp.mix+point.mix"){
      out.list <- c(out.list,
                    list(eta.grid = eta.grid,
                     pi.star = temp.ebnv$pi.star,
                     pi.tilde = temp.ebnv$pi.tilde))
    }  
    
    return(out.list)
}
```




# `hEBMF` on toy data `X2n_a` {.tabset .tabset-fade}
```{r}
set.r <- 1
set.X <- X2n_a
```

`rand.init=TRUE` is used to randomize initialization of the algorithm by permuting the eigenvalues.

## `exp`

```{r, fig.width=10, fig.height=3}
fit1 <- hEBMF(X=set.X, r=set.r, G='exp', fig.intermediate = TRUE)
fit1 <- hEBMF(X=set.X, r=set.r, G='exp', fig.intermediate = TRUE, rand.init=TRUE)
```

## `exp.mix`

```{r, fig.width=10, fig.height=3}
fit2 <- hEBMF(X=set.X, r=set.r, G='exp.mix', fig.intermediate = TRUE)
fit2 <- hEBMF(X=set.X, r=set.r, G='exp.mix', fig.intermediate = TRUE, rand.init=TRUE) 
```

## `point.mix`

```{r, fig.width=10, fig.height=3}
fit3 <- hEBMF(X=set.X, r=set.r, G='point.mix', fig.intermediate=TRUE)
fit3 <- hEBMF(X=set.X, r=set.r, G='point.mix', fig.intermediate=TRUE, rand.init=TRUE)
```

## `exp.mix+point.mix`

```{r, fig.width=10, fig.height=3}
fit4 <- hEBMF(X=set.X, r=set.r, G='exp.mix+point.mix', fig.intermediate=TRUE)
fit4 <- hEBMF(X=set.X, r=set.r, G='exp.mix+point.mix', fig.intermediate=TRUE, rand.init=TRUE)
```



#




# `hEBMF` on toy data `X2n_b` {.tabset .tabset-fade}
```{r}
set.r <- 1
set.X <- X2n_b
```

`rand.init=TRUE` is used to randomize initialization of the algorithm by permuting the eigenvalues.


## `exp`

```{r, fig.width=10, fig.height=3}
fit1 <- hEBMF(X=set.X, r=set.r, G='exp', fig.intermediate = TRUE)
fit1 <- hEBMF(X=set.X, r=set.r, G='exp', fig.intermediate = TRUE, rand.init=TRUE)
```


## `exp.mix`

```{r, fig.width=10, fig.height=3}
fit2 <- hEBMF(X=set.X, r=set.r, G='exp.mix', fig.intermediate = TRUE)
fit2 <- hEBMF(X=set.X, r=set.r, G='exp.mix', fig.intermediate = TRUE, rand.init=TRUE)
```

## `point.mix`

```{r, fig.width=10, fig.height=3}
fit3 <- hEBMF(X=set.X, r=set.r, G='point.mix', fig.intermediate=TRUE)
fit3 <- hEBMF(X=set.X, r=set.r, G='point.mix', fig.intermediate=TRUE, rand.init=TRUE)
```

## `exp.mix+point.mix`

```{r, fig.width=10, fig.height=3}
fit4 <- hEBMF(X=set.X, r=set.r, G='exp.mix+point.mix', fig.intermediate=TRUE)
fit4 <- hEBMF(X=set.X, r=set.r, G='exp.mix+point.mix', fig.intermediate=TRUE, rand.init=TRUE)
```




#




# `hEBMF` on toy data `X2` {.tabset .tabset-fade}
```{r}
set.r <- 2
set.X <- X2
```

## `exp`

```{r, fig.width=10, fig.height=3}
fit1 <- hEBMF(X=set.X, r=set.r, G='exp', fig.intermediate = TRUE)
```

## `exp.mix`

```{r, fig.width=10, fig.height=3}
fit2 <- hEBMF(X=set.X, r=set.r, G='exp.mix', fig.intermediate = TRUE)
```

## `point.mix`

```{r, fig.width=10, fig.height=3}
fit3 <- hEBMF(X=set.X, r=set.r, G='point.mix', fig.intermediate=TRUE)
```

## `exp.mix+point.mix`

```{r, fig.width=10, fig.height=3}
fit4 <- hEBMF(X=set.X, r=set.r, G='exp.mix+point.mix', fig.intermediate=TRUE)
```

#







# More toy data {.tabset .tabset-fade}

```{r, fig.width=10, fig.height=2}
more.toy <- function(seed.no){
  
    set.seed(seed.no)
  
    # create X
    p = 1000
    n = 20
    f = list()
    for(i in 1:6){ 
      f[[i]] = rnorm(p)
    }
    X =matrix(0,ncol=4*n, nrow=p)
    X[,1:(2*n)] = f[[1]]
    X[,(2*n+1):(4*n)] = f[[2]]
    
    X[,1:n] = X[,1:n]+f[[3]]
    X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
    X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
    X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
    X.svd = svd(X)
    
    # create X2
    X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

    # take transpose
    X2 <- t(X2)
    
    mat.X <- X2 + rnorm(4*20*1000, sd=3)
    
    
    obj.fit1 <-  hEBMF(X=mat.X, r=1, G='exp.mix', fig.final=FALSE)
    obj.fit2 <-  hEBMF(X=mat.X, r=1, G='point.mix', fig.final=FALSE)
    obj.fit3 <-  hEBMF(X=mat.X, r=1, G='exp.mix+point.mix', fig.final=FALSE)
    
    par(mfrow=c(1,4))
    plot(obj.fit1$l.eigvector, main='eigvector', ylab='eigenvector')
    plot(obj.fit1$l.eigvector, obj.fit1$l.map, main='vs. exp.mix', ylab='map', xlab='eigvector'); 
      abline(a=0, b=1); abline(a=0, b=-1); abline(a=0, b=0)
    plot(obj.fit2$l.eigvector, obj.fit2$l.map, main='vs. point.mix', ylab='map', xlab='eigvector'); 
      abline(a=0, b=1); abline(a=0, b=-1); abline(a=0, b=0)
    plot(obj.fit3$l.eigvector, obj.fit3$l.map, main='vs. exp.mix+point.mix', ylab='map', xlab='eigvector'); 
      abline(a=0, b=1); abline(a=0, b=-1); abline(a=0, b=0)
    
}
```

## `773`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=773)
```

## `312`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=312)
```

## `1`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=1)
```

## `2`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=2)
```

## `3`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=3)
```

## `4`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=4)
```

## `5`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=5)
```

## `6`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=6)
```

## `7`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=7)
```

## `8`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=8)
```

## `9`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=9)
```

## `10`

```{r, fig.width=10, fig.height=3}
more.toy(seed.no=10)
```

#

