---
title: "Likelihood Concentration: Random Matrix"
author: "Joonsuk Kang"
date: "2020-10-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We check the theoretical and simulation results for the probability concentration under the scenario 3: random matrix.

### Settings

\[
Z=(z_1,\dots,z_N)^T \sim N(0,\Sigma=\sigma^2(\frac{1}{\delta}+r^2)(kI_N-XX^T)^{-1}).
\]


$k^*$ is chosen to satisfy the moment condition:
\[
E[Z^TZ] =\sum_i^N \frac{\sigma^2(\frac{1}{\delta}+r^2)}{k^*-d_i}.
\]

Then, we compute the variance:
\[
var(Z^TZ) =2\sum_i^N \bigg( \frac{\sigma^2(\frac{1}{\delta}+r^2)}{k^*-d_i}  \bigg)^2.
\]



## Simulation

```{r}
library(mvtnorm)
library(ggplot2)
```


```{r} 
# define function to run simulation
sim <- function(y, # constant N/P ratio
                P, # number of columns of X
                sig2, # sigma^2: variance of E_{ij}
                delta, # delta: variance of f_j = delta*sigma^2
                r2,  # r^2
                densityplot_eigen=FALSE,
                histogram_ztz=FALSE){

N <- y*P
target <- r2/(sig2*(1/delta+r2))

# generate data
X <- matrix(data=c(rnorm(N*P, mean=0, sd=1/sqrt(P))), ncol=P, nrow=N)
Xt <- X%*%t(X) # N x N matrix
d <- eigen(Xt)$values
if(densityplot_eigen==TRUE){
    fig1 <- ggplot()+
      geom_density(aes(x=d))+
      geom_jitter(aes(d,0), height = 0.01)+
      ggtitle(expression(paste("Density of eigenvalues ",d[i])))
    print(fig1)
}
  
cond1m <- function(k,d){return(sum(1/(k-d)))} # first moment condition
loss1m <- function(k){return((cond1m(k,d)-target)^2)}
kstar <- optimize(f=loss1m, 
                   interval=c(d[1]+1/target,d[1]+N/target)
                   # min: d1+1/target (when d1 dominates others)
                   # max: d1+N/target (when d1=...=dN)
)$minimum
varztz <- 2*(sig2*(1/delta+r2))^2*sum(1/(kstar-d)^2)

# generate z from dist
z <- rmvnorm(n=1000, 
        mean=rep(0,N),
        sigma=sig2*(1/delta+r2)*solve((kstar*diag(N)-Xt))
)
ztz <- rowSums(z^2)

if(histogram_ztz==TRUE){
  fig2 <- ggplot()+geom_histogram(aes(x=ztz), bins=50)+
    ggtitle(expression(paste("Simulated Distribution of ", Z^T, Z)))
  print(fig2)
}

df.result <- 
  data.frame(theoretical.mean = r2, # theoretical mean and variance of the norm of simulated vectors Z
           theoretical.var = varztz,
           simulated.mean = mean(ztz), # mean and variance 
           simulated.var = sd(ztz)^2)
return(df.result)
}

```

The histogram of $Z^TZ$ for two cases when $P=100$ and $P=10000$ are shown.

```{r}
df.sim <- data.frame(P=c(100,200,500,1000,2000,5000,10000),
                     theoretical.mean = 0,
                     theoretical.var = 0,
                     simulated.mean = 0,
                     simulated.var = 0
                     )

for (i in 1:nrow(df.sim)){
  df.sim[i,2:5] <- sim(y = 0.3,  P = df.sim$P[i],    sig2 = 1,   delta = 1,   r2 = 1,
                       histogram_ztz=((i==1) + (i==nrow(df.sim))))
}
```

## Results


```{r}
ggplot(data=df.sim)+
  geom_point(aes(x=P, y=simulated.var, col="simulated variance"))+
  geom_line(aes(x=P, y=simulated.var, col="simulated variance"))+
  geom_point(aes(x=P, y=theoretical.var, col="theoretical variance"))+
  geom_line(aes(x=P, y=theoretical.var, col="theoretical variance"))+
  scale_x_log10()+
  scale_y_log10()+
  ylab('variance')

print(df.sim)
```
