---
title: "The Rank-1 Problem"
author: "Joonsuk Kang"
date: "2020-11-3"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


We introduce the rank-1 problem we try to solve using the `integrate out f` approach. The data and previous analysis outlined in  https://stephens999.github.io/misc/tree_pca_03.html is restated and a few additional analyses are added.


## Data from previous investigation

### X: baseline

```{r}
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]

X.svd = svd(X)

# noisy obs: Xn_a (hard) and Xn_b (easy)
set.seed(9) # the seed where flash didn't work
Xn_a = X + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
Xn_b = X + rnorm(4*n*p,sd=3)
```

### X2: simpler version

Previous analysis: "remove the effects of the top branches so we are just left with the “difficult” part of the problem"

```{r}
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])
```

### X2n_a/b: simpler version + noises 


```{r}
# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
```



## Fit from flash and svd

### fitting

```{r}
X.svd= svd(X);        X2.svd = svd(X2)
Xn_a.svd = svd(Xn_a); X2n_a.svd = svd(X2n_a)
Xn_b.svd = svd(Xn_b); X2n_b.svd = svd(X2n_b)

library(tidyverse)
library(flashr)

X.flash = flashr::flash(X,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
Xn_a.flash = flashr::flash(Xn_a,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
Xn_b.flash = flashr::flash(Xn_b,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)

X2.flash = flashr::flash(X2,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_a.flash = flashr::flash(X2n_a,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_b.flash = flashr::flash(X2n_b,10,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
```


### estimated `l`: four-factor case

```{r}
# plots
data.frame(idx = 1:80,
           l = c(X.svd$v[,1:4],
                 Xn_a.svd$v[,1:4],
                 Xn_b.svd$v[,1:4],
                 c(X.flash$ldf$f[,1:4]),
                 c(Xn_a.flash$ldf$f[,1:4]),
                 c(Xn_b.flash$ldf$f[,1:4])),
           factor=rep(rep(1:4, each=80),times=6),
           data = rep(rep(c("X", "Xn_a (hard)", "Xn_b (easy)"), each=4*80), times=2),
           method = rep(c("svd", "flash"), each=80*4*3)
           )  %>% 
  # adjust sign for illustration
  mutate(l = ifelse((factor==3 & data=="X" & method=="flash")|
                              (factor==4 & data=="Xn_a (hard)" & method=="flash")|
                              (factor==4 & data=="Xn_b (easy)" & method=="flash"),
                            -l,
                            l)
         ) -> X.fit
X.fit %>%
  ggplot()+geom_line(aes(x=idx, y=l, group=method, col=method))+
  facet_grid(data~factor, labeller=label_both)+geom_hline(yintercept=0)+
  ggtitle("Four factors")
```

The flash fits are identical to the svd results (singular vectors) in all cases. 

The first factor captures the mean trend and the second factor captures the (-1,-1,1,1) split. The result is consistent across data (noiseless `X` and noisy `Xn_a`/`Xn_b`) and method (flash and svd).

For the third and fourth factor, we want to obtain a sparse solution, such as factor 3=(0,0,-1,1) and factor 4 = (-1,1,0,0). The noiseless data `X` and the noisy data `Xn_a` have the nonsparse solution while the noisy data `Xn_b` has the sparse solution. 

To focus on the factors 3 and 4, we constructed a simpler data `X2` and its corresponding noisy versions `X2n_a` and `X2n_b`.



### estimated `l`: simplified two-factor case

```{r}
data.frame(idx = 1:80,
           l = c(X2.svd$v[,1:2],
                 X2n_a.svd$v[,1:2],
                 X2n_b.svd$v[,1:2],
                 c(X2.flash$ldf$f[,1:2]),
                 c(X2n_a.flash$ldf$f[,1:2]),
                 c(X2n_b.flash$ldf$f[,1:2])),
           factor=rep(rep(1:2, each=80),times=6),
           data = rep(rep(c("X2", "X2n_a (hard)", "X2n_b (easy)"), each=2*80), times=2),
           method = rep(c("svd", "flash"), each=80*2*3)
           ) %>% 
  # adjust sign for illustration
  mutate(l = ifelse((factor==1 & data=="X2n_b (easy)" & method=="flash"),
                            -l,
                            l)
         ) -> X2.fit

# warmstart: sparse fit (from X2n_b) initialization for flash on data X2
X2.flash.warmstart = flashr::flash(X2,K=4,f_init=X2n_b.flash,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant",backfit = TRUE,greedy = FALSE, verbose=FALSE)
data.frame(idx=1:80, 
           l = c(X2.flash.warmstart$ldf$f[,1:2]), 
           factor=rep(1:2, each=80), data="X2", method="flash_warm") %>%
  rbind(X2.fit) -> X2.fit

X2.fit %>%
  ggplot()+geom_line(aes(x=idx, y=l, group=method, col=method))+
  facet_grid(data~factor, labeller=label_both)+#geom_hline(yintercept=0)+
  ggtitle("Simplified two factors")
```

As expected, the results are identical to the previous result for the third and fourth factors.

The warmstart fit is added. It uses the sparse fit from `X2n_b` as the initialization for the flash model on data `X2`. The wartmstart fit produces an enhanced sparse solution.


## The rank-1 problem

The preceding problem was to find both of the two factors in the data. Now, we further simplify it to a rank-1 problem.

**The rank-1 problem** We are given the same set of data: noiseless rank-2 matrix `X2`, noisy versions `X2n_a` and `X2n_b`. And we are allowed to fit only one factor. The goal is to find a solution that is sparse.

As shown in https://joonsukkang.github.io/mf/warmstart.html, the warmstart fit reverts to the nonsparse solution (which is identical to svd or flash fit) when we only allow rank-1 estimation. So, under this rank-1 problem setting, warmstart does not give you an advantage. 

```{r}
# flash fit: restricted to rank-1
X2.flash1 = flashr::flash(X2,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_a.flash1 = flashr::flash(X2n_a,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)
X2n_b.flash1 = flashr::flash(X2n_b,1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant", verbose=FALSE)

# 
data.frame(idx = 1:80,
           l = c(X2.svd$v[,1],
                 X2n_a.svd$v[,1],
                 X2n_b.svd$v[,1],
                 X2.flash1$ldf$f[,1],
                 X2n_a.flash1$ldf$f[,1],
                 X2n_b.flash1$ldf$f[,1]),
           data = rep(rep(c("X2", "X2n_a (hard)", "X2n_b (easy)"), each=80), times=2),
           method = rep(c("svd", "flash"), each=80*3)
           ) %>% 
  # adjust sign for illustration
  mutate(l = ifelse((data=="X2n_b (easy)" & method=="flash"),
                            -l,
                            l)
         ) -> X2.fit1

# warmstart: sparse fit (from X2n_b) initialization for flash on data X2
X2.flash.warmstart1 = flashr::flash(X2,K=1,f_init=X2n_b.flash1,ebnm_fn = list(l="ebnm_pn", f="ebnm_ash"),var_type = "constant",backfit = TRUE,greedy = FALSE, verbose=FALSE)
data.frame(idx=1:80, 
           l = c(X2.flash.warmstart1$ldf$f[,1]), data="X2", method="flash_warm") %>%
  rbind(X2.fit1) -> X2.fit1

X2.fit1 %>%
  ggplot()+geom_line(aes(x=idx, y=l, group=method, col=method))+
  facet_grid(method~data, labeller=label_both)+geom_hline(yintercept=0)+
  ggtitle("The Rank-1 Problem")
```
