---
title: "toy data from Yusha: nonneg init (original data)"
author: "Joonsuk Kang"
date: "2021-10-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# obeservations

* Note: here, I am using the original version of Yusha's toy data (not F1 scaled)

* alt-flash with a nonnegative initialization using eigenvectors (implemented below) leads to a more interpretable result (the patterns more kept) than alt-flash with the oracle initialization

* in terms of elbo, oracle init tends to achieves a higher elbo (in experiments with seed=1,2,4,5; nonneg init's elbo was higher in when seed=3)





# some functions

### Initialize L (direction) for nonnegative priors

the first (root) factor takes the first eigenvector (replace negative values with 0, if any)

add factors by taking eigenvector of the residual matrix (the parts not explained by the existing factors) and replace negative values with 0.

stop adding factors if the quadratic programming solution for finding optimal scale has the min/max ratio less than a threshold (1e-4).

```{r}
# return nonnegative direction estiamtes (L)
INIT.L.NonNeg <- function(Y, verbose=1, max.K=20, tol.scale.ratio=1e-10){

  n <- nrow(Y)

  # residual Y
  Yres <- Y

  # root factor
  u <- eigen(Yres)$vectors[,1]
  u <- u * sign(u[which.max(abs(u))])
  u[u<0] <- 0
  u <- u/sqrt(sum(u^2))
  L <- matrix(u, ncol=1)
  d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
  Yres <- Y - L %*% diag(d, nrow=length(d)) %*% t(L)

  # add factors
  for (k in 2:max.K){
    ## find direction
    u <- eigen(Yres)$vectors[,1]
    u <- u * sign(u[which.max(abs(u))])
    u[u<0] <- 0
    u <- u/sqrt(sum(u^2))

    ## find scale
    L <- cbind(L, matrix(u, ncol=1))
    d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
    if(verbose==1){print(paste0(k,'-th factor fitting; ', 'min/max sacle ratio: ', min(d)/max(d)))}
    if(min(d)/max(d) < tol.scale.ratio) {
      L <- L[,-k]
      d <- scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$scale
      break
    }
    Yres <- Y - L %*% diag(d, nrow=length(d)) %*% t(L)
  }

  if(verbose==1){print(paste0(ncol(L), ' factors fitted'))}
  out.list <- list(L=L)
  return(out.list)
}

```

### some more functions

```{r}
library(tidyverse)
library(gridExtra)

# plot posterior mean of L and F 
plotLF <- function(A.l, A.f, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             F=c(A.f),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figL
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=F, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figF
  
  grid.arrange(figL, figF, nrow=2, 
               top=title)
}

plotL <- function(A.l, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') +ggtitle(title)-> figL
  
  print(figL)
}


# scale L: quadratic programming approach to find optimal scaling
scaleL.qp <- function(Y, L, scale.min, print.scale=TRUE){

  n <- nrow(L); K <- ncol(L)
  Ls <- matrix(0, nrow=n, ncol=K)
  
  # normalized L: column norm 1
  Ln <- L %*% diag(x=1/sqrt(colSums(L^2)),nrow=K)
  
  D <- crossprod(Ln)^2
  d <- rep(0, K);   for (k in 1:K){d[k] <- as.numeric(t(Ln[,k])%*%Y%*%Ln[,k])}
  qp <- quadprog::solve.QP(Dmat=D, dvec=d, Amat=diag(1,nrow=K), bvec=rep(scale.min,K)) 
  
  scale <- qp$solution
  if(print.scale){print(scale)}

  for (k in 1:K){Ls[,k] <- sqrt(scale[k])*Ln[,k]}
  
  out.list <- list(scale=scale, Ls=Ls)
  
  return(out.list)
}
```


# results

```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211017.R")
```

```{r}
# source: https://yushaliu.github.io/single_cell_cancer/simulations3_update.html
model <- readRDS(url("https://raw.githubusercontent.com/YushaLiu/single_cell_cancer/master/data/model.rds"))
L <- model$L
F <- model$F
F[,1] <- F[,1] - 10 # center the data
```

```{r}
set.seed(1)
X <- L%*%t(F) + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X%*%t(X)
Y <- XXt
```

### nonnegative initialization (using eigenvectors of Y)

```{r}
temp.obj <- INIT.L.NonNeg(Y, verbose=1) # initialize directions (L) 
Ls <- scaleL.qp(Y, temp.obj$L, scale.min=1e-10, print.scale=FALSE)
plotL(Ls$Ls, title='nonneg initialized L')

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: alt-flash with nonneg init')
```

### oracle initialization

```{r}
Ls <- scaleL.qp(Y, L, scale.min=1e-10)
plotL(Ls$Ls, title='oracle L') 

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
print(fit.alt$scales)
plotLF(fit.alt$A.l, fit.alt$A.f, 'result: alt-flash with oracle init')
```


### some more seeds

since the fitted L and F are identical, I report only L

```{r}
for (seed in 2:5){
  
  set.seed(seed)
  X <- L%*%t(F) + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
  Y <- X %*% t(X)
  
  print(paste0('seed=',seed))
  
  # nonneg init
  temp.obj <- INIT.L.NonNeg(Y, verbose=0)
  Ls <- scaleL.qp(Y, temp.obj$L, scale.min=1e-10, print.scale=FALSE)
  plotL(Ls$Ls, title=paste0('seed=',seed, ': nonneg initialized L'))

  fit.alt <- alt.flash(INIT.L(init.L=Ls$Ls, Y=Y))
  print(fit.alt$scales)
  plotL(fit.alt$A.l, paste0('seed=',seed,': alt-flash w/ nonneg init'))
  
  # oracle init
  fit.alt <- alt.flash(INIT.L(init.L=scaleL.qp(Y, L, scale.min=1e-10, print.scale=FALSE)$Ls, Y=Y))
  print(fit.alt$scales)
  plotL(fit.alt$A.l, paste0('seed=',seed,': alt-flash w/ oracle init'))
}
```

