---
title: "toy data from Yusha: oracle initialization "
author: "Joonsuk Kang"
date: "2021-10-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# summary

* the first root factor $F_1$ seems to be too strong to allow other factors to be meaningfully captured

* if we cut back the scale of $F_1$ by half, alt-flash returns an estimated $L$ (=$F$) that captures all 11 factors (with some noise)


# some functions for investigation

```{r}
library(tidyverse)
library(gridExtra)

# plot posterior mean of L and F 
plotLF <- function(A.l, A.f, title){

  K <- ncol(A.l)
  n <- nrow(A.l)
  
  data.frame(L=c(A.l),
             F=c(A.f),
             factor=rep(1:K, each=n),
             x=rep(1:n, times=K)) %>%
    mutate(factor=as.factor(factor)) -> df.plot
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figL
  
  df.plot %>%
    ggplot()+
    geom_line(aes(x=x, y=F, group=factor, col=factor))+
    facet_wrap(~factor, scales='free_y') -> figF
  
  grid.arrange(figL, figF, nrow=2, 
               top=title)
}


# scale L: quadratic programming approach to find optimal scaling
scaleL.qp <- function(Y, L, scale.min, print.scale=TRUE){

  n <- nrow(L); K <- ncol(L)
  Ls <- matrix(0, nrow=n, ncol=K)
  
  # normalized L: column norm 1
  Ln <- L %*% diag(x=1/sqrt(colSums(L^2)))
  
  D <- crossprod(Ln)^2
  d <- rep(0, K);   for (k in 1:K){d[k] <- as.numeric(t(Ln[,k])%*%Y%*%Ln[,k])}
  qp <- quadprog::solve.QP(Dmat=D, dvec=d, Amat=diag(1,nrow=K), bvec=rep(scale.min,K)) 
  
  scale <- qp$solution
  if(print.scale){print(scale)}

  for (k in 1:K){Ls[,k] <- sqrt(scale[k])*Ln[,k]}
  
  out.list <- list(scale=scale, Ls=Ls)
  
  return(out.list)
}
```




# toy data: original version

```{r}
# source: https://yushaliu.github.io/single_cell_cancer/simulations3_update.html

library(Matrix)
library(flashier)
library(magrittr)
library(RColorBrewer)
library(pheatmap)


source(url("https://raw.githubusercontent.com/YushaLiu/single_cell_cancer/master/code/util.R"))

### load in simulated loadings and factors
model <- readRDS(url("https://raw.githubusercontent.com/YushaLiu/single_cell_cancer/master/data/model.rds"))
L <- model$L
F <- model$F
F[,1] <- F[,1] - 10 # center the data

### plot the drift factorization of truth, where F is orthogonal across columns 
pheatmap(L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True L")
pheatmap(F, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")
pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")
```


# preliminary result (unsatisfactory); sd=0.5

From $XX^T=(LF^T+E)(LF^T+E)^T\approx LF^TFL^T$, use the diagonal of $F^TF$ as the scaling factor.

```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211017.R")
```

```{r}
set.seed(1)

X <- L%*%t(F) + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X%*%t(X)
Y <- XXt
Ls <- L %*% diag(sqrt(as.vector(diag(t(F)%*%F))))

fit.init <- INIT.L(init.L=Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
plotLF(fit.alt$A.l, fit.alt$A.f, 'preliminary result')
```



# looking into data

The Quadratic Programming approach: 

To assess the difficulty (or well-definedness) of the problem, we can check the optimal weights $d_1,\dots,d_K$ that minimizes $\|Y-\sum_k d_k l_kl_k^T\|_F^22$ where $Y$ is the $XX^T$ matrix and $l_k$ is the $k$-th vector of the true matrix $L$, scaled to norm 1, and $(d_1,\dots,d_K)$ is the scale vector. 


### noiseless case: approximating $LF^TFL^T$  

```{r}
# the first column is too powerful!
sqrt(colSums(F^2))

X0 <- L %*% t(F) 
X0X0t <- X0 %*% t(X0)
Ls <- scaleL.qp(Y=X0X0t, L, scale.min=0, print.scale=FALSE)

```

The solution is 
```{r}
print(Ls$scale)
```
and in a relative scale compared to the largest one ($d_1$), the scales are
```{r}
print(Ls$scale/max(Ls$scale))
```
Notice that the smallest one ($d_4$) is 3.173557e-05 times the size of the largest one.

So, even for a noiseless case, given all $K$ true bases ($L$), the problem is close to ill-defined.


### noisy cases: with noise level sd=0.5

The noisy case is difficult as well, but not necessarily more difficult than the noiseless one.

*The difficulty comes from the strong first factor ($F_1$), not from high noise level.*

```{r}
out <- rep(0,100)
for (seed in 1:length(out)){
  set.seed(seed)
  X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
  XXt <- X %*% t(X)
  Ls <- scaleL.qp(Y=XXt, L, scale.min=0, print.scale=FALSE)
  out[seed] <- min(Ls$scale)/max(Ls$scale)
}
summary(out)
```



# what if: the first root factor $F_1$ has smaller effect

Let's reduce the magnitude of the first factor by half.

```{r}
F[,1] <- F[,1]/2
```


For the noiseless data, the smallest one ($d_4$) is now 0.001624 compared to the largest; it is 50 times larger than before (3.173557e-05).

```{r}
X0 <- L %*% t(F) 
X0X0t <- X0%*%t(X0)
Ls <- scaleL.qp(Y=X0X0t, L, scale.min=0)
min(Ls$scale)/max(Ls$scale)
```

Also, for the noisy data with sd=0.5, the ratio is on the same magnitude with the noiseless one.

```{r}
out <- rep(0,100)
for (seed in 1:length(out)){
  set.seed(seed)
  X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
  XXt <- X %*% t(X)
  Ls <- scaleL.qp(Y=XXt, L, scale.min=0, print.scale=FALSE)
  out[seed] <- min(Ls$scale)/max(Ls$scale)
}
summary(out)
```

# now, somewhat surprisingly, we can solve the problem

Remember: the $F$ used below is the one after rescaling the first factor by half.

```{r}
set.seed(1)
X0 <- L %*% t(F) 
X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
XXt <- X %*% t(X)
```


# results

### `F^TF init` (using true $L$ and $F$)

From $XX^T=(LF^T+E)(LF^T+E)^T\approx LF^TFL^T$, use the diagonal of $F^TF$ as the scaling factor.

```{r}
Y <- XXt
Ls <- L %*% diag(sqrt(as.vector(diag(t(F)%*%F))))

fit.init <- INIT.L(init.L=Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
plotLF(fit.alt$A.l, fit.alt$A.f, 'FtF initialization')
```


### `qp init` (using true $L$)

Use the optimal scales from quadratic programming approach.

```{r}
Y <- XXt
Ls <- scaleL.qp(Y, L, scale.min=0)

fit.init <- INIT.L(init.L=Ls$Ls, Y=Y)
fit.alt <- alt.flash(fit.init)
plotLF(fit.alt$A.l, fit.alt$A.f, 'qp initialization')
```


### some more seeds

```{r}
for (seed in 2:5){
  
  set.seed(seed)
  X <- X0 + matrix(rnorm(800*2000, 0, 0.5), nrow=800, ncol=2000)
  Y <- X %*% t(X)
  
  print(paste0('seed=',seed))
  
  # F^TF init
  fit.alt <- alt.flash(INIT.L(init.L=L %*% diag(sqrt(as.vector(diag(t(F)%*%F)))), Y=Y))
  plotLF(fit.alt$A.l, fit.alt$A.f, paste0('seed=',seed,': FtF initialization'))
  
  # qp init
  fit.alt <- alt.flash(INIT.L(init.L=scaleL.qp(Y, L, scale.min=0, print.scale=FALSE)$Ls, Y=Y))
  plotLF(fit.alt$A.l, fit.alt$A.f, paste0('seed=',seed,': qp initialization'))
}
```



