---
title: "Tree Toy Data: flash results w/ denoising"
author: "Joonsuk Kang"
date: "2021-10-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(flashier)
```


# Summary

- the goal is to extract the `(1,0,0,0), (0,1,0,0), (0,0,1,0), (0,0,0,1)` pattern 
and the method used is flshier on `cov(tX))` with point exponential prior

- here we implement the rank-2 denoising idea, which greatly improves the fit

- the toy data have rank-2 plus noise structure, which can be easily detected using a scree plot

- we can recover three patterns from `X2` and `X2n_a`, and two from `X2n_b`; notice that now `X2n_b` is a harder problem

- the denoising idea greatly improves the fit, though flashier is still missing one or two of the patterns.

- backfit does not seem to improve fit significantly







# Toy data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

rm(X, X.svd, f, i, n, p)
```

The toy data have a rank-2 plus noise structure (we know since we made them). 
And as we can see, the singular values have two large values.
We create a rank-2 version of the toy data to check whether denoising before model fitting
greatly improves the model fit.


```{r}
plot(svd(X2)$d)
plot(svd(X2n_a)$d)
plot(svd(X2n_b)$d)

# create a rank-2 denoised version of matrix
r2d <- function(dat.mat){
  smat <- svd(dat.mat)
  dmat <- smat$u[,1:2] %*% diag(smat$d[1:2], nrow=2) %*% t(smat$v[,1:2])
  return(dmat)
}
dX2 <-r2d(X2)
dX2n_a <-r2d(X2n_a)
dX2n_b <-r2d(X2n_b)
```

# Some Functions

```{r}
# a function for plotting
plot.three <- function(L.X2, L.X2n_a, L.X2n_b){
  data.frame(L=c(c(L.X2), c(L.X2n_a), c(L.X2n_b)),
             x=1:80, 
             factor =rep(as.factor(c(1,2)), each=80),
             data = rep(c('X2', 'X2n_a', 'X2n_b'), each=80*2)) %>%
    ggplot()+geom_line(aes(x=x, y=L, col=factor))+facet_wrap(~data)
}
```


```{r}
# add 'point_exponential' to the choice of flashier priors
prior.point.exponential <- function(...) {
  args <- as.prior.args(prior.family = "point_exponential",
                        optmethod = "nlm",
                        ...)
  return(do.call(as.prior, args))
}
environment(prior.point.exponential) <- asNamespace('flashier') 

plot.factors <- function(fit, showK=NULL){
  if(is.null(showK)){showK <- ncol(fit$loadings.pm[[1]])}
  
  data.frame(L=c(fit$loadings.pm[[1]][,1:showK]), 
           x=rep(1:80),
           factor=as.factor(rep(1:showK, 
                      each=80))) %>%
  ggplot()+geom_line(aes(x=x, y=L, col=factor))
}
```

# Results

## data `X2`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(dX2)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
plot.factors(fit1)

# with backfit
fit1b <- flash(cov(t(dX2)), verbose.lvl=0, 
              backfit=TRUE, prior.family = prior.point.exponential())
plot.factors(fit1b)
```

## data `X2n_a`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(dX2n_a)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
plot.factors(fit1)

# with backfit
fit1b <- flash(cov(t(dX2n_a)), verbose.lvl=0,
              backfit=TRUE, prior.family = prior.point.exponential())
plot.factors(fit1)
```

## data `X2n_b`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(dX2n_b)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
plot.factors(fit1)

# with backfit
fit1b <- flash(cov(t(dX2n_b)), verbose.lvl=0, 
              backfit=TRUE, prior.family = prior.point.exponential())
plot.factors(fit1)
```
