---
title: "Tree Toy Data: alt-flash results (1)"
author: "Joonsuk Kang"
date: "2021-10-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(flashier)
library(ebnm)
prior.point.exponential <- function(...) {
  args <- as.prior.args(prior.family = "point_exponential",
                        optmethod = "nlm",
                        ...)
  return(do.call(as.prior, args))
}
environment(prior.point.exponential) <- asNamespace('flashier') 
```

# Toy data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

rm(X, X.svd, f, i, n, p)
```


# Method

In this investigation, we use `alt-flash` with `point-exponential` prior. 

The initialization seems to play an important role in the model fitting in that 
we need to provide an initial condition for parameters that falls in a feasible reason 
for the alt-flash algorithm to work properly.

I tried a two-step fitting procedure that 

- the first step involves finding a stable state given priors $g_l$ and $g_f$. 
To be specific, we generate initial $L$ from $g_l$ and $F$ from $g_f$, and then
update $q_l, q_f, \tau$ with the priors $g_l,g_f$ fixed, until the ELBO stabilizes.

- the second step is the full backfitting procedure (now start updating $g_l$ and $g_f$ as well). 


This sensitivity to initial conditions does not necessarily mean a sensitivity 
to the prior choices $g_l$ and $g_f$, which needs further investigation.


In the following investigation, we use the following simple point exponential prior 
as the initial values of `g.l` and `g.f`.

```{r}
# a point-exponential prior
g.pe <- gammamix(pi=c(0.1, 0.9), shape=c(1,1), scale=c(0,1))
```


```{r}
setwd("~/Box/2-research/mf")
source("code/alt-flash_v20211010.R")
```


```{r}
list.to.elbo <- function(out.list){
  
  df <- data.frame(elbo=double(),
                   iteration=integer(),
                   K=integer(),
                   seed=integer())

  for (K in 1:length(out.list)){
    for (seed in 1:length(out.list[[K]])){
      df <- rbind(df, 
                  data.frame(elbo=out.list[[K]][[seed]]$elbo.vec,
                             iteration = 1:length(out.list[[K]][[seed]]$elbo.vec),
                             K=K,
                             seed=seed))
    }
  }
  return(df)
}


list.to.factors <- function(out.list){
  
  df <- data.frame(L = double(),
                   x = integer(),
                   factor = integer(),
                   K = integer(),
                   seed = integer()
                   )

  for (K in 1:length(out.list)){
    for (seed in 1:length(out.list[[K]])){
      
      mat.temp <- out.list[[K]][[seed]]$A.l # use posterior mean of L
      mat.temp <- mat.temp %*% diag(1/sqrt(colSums(mat.temp^2)),  # scale columns to L2 norm 1
                                    nrow=ncol(mat.temp),
                                    ncol=ncol(mat.temp))
      
      df <- rbind(df, 
                  data.frame(L = c(mat.temp),
                             x = 1:nrow(mat.temp),
                             factor = rep(1:ncol(mat.temp), each=nrow(mat.temp)),
                             K = K,
                             seed = seed))
    }
  }
  return(df)
}

top10 <- function(df.elbo, df.L){
	
	df.elbo %>% 
    group_by(K, seed) %>%
    summarise(elbo=last(elbo)) %>%
    ungroup() %>%
    arrange(-elbo) %>% head(10) %>%
    mutate(rank=1:10) %>% select(rank, elbo, K, seed) -> df.elbo.top10

  print(df.elbo.top10)
  
  df.elbo.top10 %>% 
    select(-elbo) %>% 
    inner_join(df.L, by=c('K', 'seed'))  %>% 
    mutate(factor=as.factor(factor)) %>%
    ggplot()+geom_line(aes(x=x, y=L, group=factor, col=factor))+
    facet_wrap(~rank, nrow=2, labeller='label_both') -> fig
  
  print(fig)
}

```


# Results: `cov(t(X2))`


```{r, eval=FALSE}
# for record: g.pe <- gammamix(pi=c(0.1, 0.9), shape=c(1,1), scale=c(0,1))
out.list <- list()
data.mat <- cov(t(X2))

for (K in 1:8){
  print(paste0("now fitting K=",K)) 
  temp.list <- list()
  g.l <- rep(list(g.pe), K)
  g.f <- rep(list(g.pe), K)
      for (seed in 1:5){ 
        temp.list[[seed]] <- alt.flash(Y=data.mat, g.l=g.l, g.f=g.f, seed=seed)
        print(paste0("K=",K, "; seed=", seed, " completed"))
      }
  out.list[[K]] <- temp.list
}

saveRDS(out.list, "output/tree_toy_altflash_X2_v2021_1012_1309.rds")
```

```{r}
results_X2 <- readRDS("output/tree_toy_altflash_X2_v2021_1012_1309.rds")

list.to.elbo(results_X2) %>%
  mutate(seed=as.factor(seed),
         dist=max(elbo)-elbo+1e-6) %>%
  ggplot()+
  geom_line(aes(x=iteration, y=dist, group=seed, col=seed)) +
  scale_y_log10()+
  facet_grid(~K, labeller=labeller(.cols=label_both))+
  ylab('distance from best elbo')

list.to.factors(results_X2) %>% 
  mutate(factor=as.factor(factor)) %>%
  ggplot()+geom_line(aes(x=x, y=L, group=factor, col=factor))+
  facet_grid(K~seed, 
             labeller = labeller(.rows = label_both, .cols = label_both))
```


Observations

- when $K>4$ is given, only four factors are used and the residual factors are dropped; the desired `(1,0,0,0), (0,1,0,0), (0,0,1,0), (0,0,0,1)` pattern is captured

- when $K<4$ is given, a sub optimal fit is achieved; capturing only $K$-many of the four patterns

- this noiseless task is easier than the noisy tasks below

- a much higher elbo is achieved compared to flashier (when given $K\geq 4$)

```{r}
max(list.to.elbo(results_X2)$elbo)
flash(cov(t(X2)), prior.family=prior.point.exponential(), backfit=TRUE, verbose.lvl=0)$elbo
```



# Results: `X2n_a`

```{r, eval=FALSE}
# for record: g.pe <- gammamix(pi=c(0.1, 0.9), shape=c(1,1), scale=c(0,1))
out.list <- list()
data.mat <- cov(t(X2n_a))

for (K in 1:8){
  print(paste0("now fitting K=",K)) 
  temp.list <- list()
  g.l <- rep(list(g.pe), K)
  g.f <- rep(list(g.pe), K)
      for (seed in 1:5){ 
        temp.list[[seed]] <- alt.flash(Y=data.mat, g.l=g.l, g.f=g.f, seed=seed)
        print(paste0("K=",K, "; seed=", seed, " completed"))
      }
  out.list[[K]] <- temp.list
}

saveRDS(out.list, "output/tree_toy_altflash_X2n_a_v2021_1012_1309.rds")
```

```{r}

results_X2n_a <- readRDS("output/tree_toy_altflash_X2n_a_v2021_1012_1309.rds")

df.elbo <- list.to.elbo(results_X2n_a)
df.elbo %>%
  mutate(seed=as.factor(seed),
         dist=max(elbo)-elbo+1e-6) %>%
  ggplot()+
  geom_line(aes(x=iteration, y=dist, group=seed, col=seed)) +
  scale_y_log10()+
  facet_grid(~K, labeller=labeller(.cols=label_both))+
  ylab('distance from best elbo')
  
df.L <- list.to.factors(results_X2n_a)
df.L %>% 
  mutate(factor=as.factor(factor)) %>%
  ggplot()+geom_line(aes(x=x, y=L, group=factor, col=factor))+
  facet_grid(K~seed, 
             labeller = labeller(.rows = label_both, .cols = label_both))

top10(df.elbo, df.L)
flash(cov(t(X2n_a)), prior.family=prior.point.exponential(), backfit=FALSE, verbose.lvl=0)$elbo # backfit fails
```


# Results: `X2n_b`


```{r, eval=FALSE}
# for record: g.pe <- gammamix(pi=c(0.1, 0.9), shape=c(1,1), scale=c(0,1))
out.list <- list()
data.mat <- cov(t(X2n_b))

for (K in 1:8){
  print(paste0("now fitting K=",K)) 
  temp.list <- list()
  g.l <- rep(list(g.pe), K)
  g.f <- rep(list(g.pe), K)
      for (seed in 1:5){ 
        temp.list[[seed]] <- alt.flash(Y=data.mat, g.l=g.l, g.f=g.f, seed=seed)
        print(paste0("K=",K, "; seed=", seed, " completed"))
      }
  out.list[[K]] <- temp.list
}

saveRDS(out.list, "output/tree_toy_altflash_X2n_b_v2021_1012_1309.rds")
```


```{r}
results_X2n_b <- readRDS("output/tree_toy_altflash_X2n_b_v2021_1012_1309.rds")

df.elbo <- list.to.elbo(results_X2n_b)
df.elbo %>%
  mutate(seed=as.factor(seed),
         dist=max(elbo)-elbo+1e-6) %>%
  ggplot()+
  geom_line(aes(x=iteration, y=dist, group=seed, col=seed)) +
  scale_y_log10()+
  facet_grid(~K, labeller=labeller(.cols=label_both))+
  ylab('distance from best elbo')
  
df.L <- list.to.factors(results_X2n_b)
df.L %>% 
  mutate(factor=as.factor(factor)) %>%
  ggplot()+geom_line(aes(x=x, y=L, group=factor, col=factor))+
  facet_grid(K~seed, 
             labeller = labeller(.rows = label_both, .cols = label_both))

top10(df.elbo, df.L)
flash(cov(t(X2n_b)), prior.family=prior.point.exponential(), backfit=TRUE, verbose.lvl=0)$elbo
```








