---
title: "Tree Toy Data: flash results"
author: "Joonsuk Kang"
date: "2021-10-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(flashier)
```


# Summary

## extracting `(1,-1,0,0), (0,0,1,-1)` pattern

- data `X2` is a noiseless version; `X2n_a` is a 'hard' (i.e. singular vectors are not sparse) noisy version and `X2n_b` an 'easy' (i.e. singular vectors are close to zero) noisy version.

- flashier on data `X` (which means original matrices `X2`, `X2n_a`, or `X2n_b`) leads to a sparser version of singular vectors, but the results are essentially identical to singular vectors

- flashier on covariance `cov(t(X))` successfully extracts the `(1,-1,0,0), (0,0,1,-1)` pattern--using either point-normal or point-laplace prior.


## extracting `(1,0,0,0), (0,1,0,0), (0,0,1,0), (0,0,0,1)` pattern

- method: flashier on `cov(t(X))` with point exponential prior 

- for noiseless `X2`, flashier finds three, missing `(0,1,0,0)`

- for hard case `X2n_a`,  49 factors (maximum number allowed) are fitted. The first factor roughly captures `(1,0,0,0)` and the second factor captures two points in the first 20 coordinates; the others are not much interpretable.

- for easy case `X2n_b`, 5 factors are fitted. The first captures `(0,0,0,1)` while each of the other four factors captures some exclusive subsets of `(1,0,0,0)` pattern.

- backfit does not seem to improve fit significantly




# Toy data

```{r}
# code originally from https://stephens999.github.io/misc/tree_pca_03.html.
#--------------------------------------------------
# create X
set.seed(123)
p = 1000
n = 20
f = list()
for(i in 1:6){ 
  f[[i]] = rnorm(p)
}
X =matrix(0,ncol=4*n, nrow=p)
X[,1:(2*n)] = f[[1]]
X[,(2*n+1):(4*n)] = f[[2]]

X[,1:n] = X[,1:n]+f[[3]]
X[,(n+1):(2*n)] = X[,(n+1):(2*n)]+f[[4]]
X[,(2*n+1):(3*n)] = X[,(2*n+1):(3*n)] + f[[5]]
X[,(3*n+1):(4*n)] = X[,(3*n+1):(4*n)] + f[[6]]
X.svd = svd(X)

# create X2
X2 = X- X.svd$u[,1:2] %*% diag(X.svd$d[1:2]) %*% t(X.svd$v[,1:2])

# adding noise
set.seed(9) # the seed where flash didn't work
X2n_a = X2 + rnorm(4*n*p,sd=3)

set.seed(5) # the seed where flash did work
X2n_b = X2 + rnorm(4*n*p,sd=3)
#--------------------------------------------------

# take transpose
X2 <- t(X2)
X2n_a <- t(X2n_a)
X2n_b <- t(X2n_b)

rm(X, X.svd, f, i, n, p)
```


```{r}
# a function for plotting
plot.three <- function(L.X2, L.X2n_a, L.X2n_b){
  data.frame(L=c(c(L.X2), c(L.X2n_a), c(L.X2n_b)),
             x=1:80, 
             factor =rep(as.factor(c(1,2)), each=80),
             data = rep(c('X2', 'X2n_a', 'X2n_b'), each=80*2)) %>%
    ggplot()+geom_line(aes(x=x, y=L, col=factor))+facet_wrap(~data)
}
```


# singular vectors

```{r}
plot.three(svd(X2)$u[,1:2], svd(X2n_a)$u[,1:2], svd(X2n_b)$u[,1:2])
```


# extracting `(1,-1,0,0), (0,0,1,-1)` 


## flashier on `X`

```{r}
# point-normal prior; without backfit
flash.fit <- function(data.mat){
  flash(data.mat, verbose.lvl=0, 
        backfit=FALSE, prior.family = prior.point.normal())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))

# point-normal prior; with backfit
flash.fit <- function(data.mat){
  flash(data.mat, verbose.lvl=0, 
        backfit=TRUE, prior.family = prior.point.normal())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))


# point-laplace prior; without backfit
flash.fit <- function(data.mat){
  flash(data.mat, verbose.lvl=0, 
        backfit=FALSE, prior.family = prior.point.laplace())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))

# point-laplace prior; with backfit
flash.fit <- function(data.mat){
  flash(data.mat, verbose.lvl=0, 
        backfit=TRUE, prior.family = prior.point.laplace())$loadings.pm[[1]][,1:2]
}
#plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b)) 
# returns error:
#     Error in (function (f, p, ..., hessian = FALSE, typsize = rep(1, length(p)),  : 
#     non-finite value supplied by 'nlm'

```

## flashier on `cov(t(X))`

```{r}
# point-normal prior; without backfit
flash.fit <- function(data.mat){
  flash(cov(t(data.mat)), verbose.lvl=0, 
        backfit=FALSE, prior.family = prior.point.normal())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))

# point-normal prior; with backfit
flash.fit <- function(data.mat){
  flash(cov(t(data.mat)), verbose.lvl=0, 
        backfit=TRUE, prior.family = prior.point.normal())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))


# point-laplace prior; without backfit
flash.fit <- function(data.mat){
  flash(cov(t(data.mat)), verbose.lvl=0, 
        backfit=FALSE, prior.family = prior.point.laplace())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b))

# point-laplace prior; with backfit
flash.fit <- function(data.mat){
  flash(cov(t(data.mat)), verbose.lvl=0, 
        backfit=TRUE, prior.family = prior.point.laplace())$loadings.pm[[1]][,1:2]
}
plot.three(flash.fit(X2), flash.fit(X2n_a), flash.fit(X2n_b)) 
```



# extracting `(1,0,0,0), (0,1,0,0), (0,0,1,0), (0,0,0,1)` 

```{r}
# add 'point_exponential' to the choice of flashier priors
prior.point.exponential <- function(...) {
  args <- as.prior.args(prior.family = "point_exponential",
                        optmethod = "nlm",
                        ...)
  return(do.call(as.prior, args))
}
environment(prior.point.exponential) <- asNamespace('flashier') 

plot.factors <- function(fit, showK){
  data.frame(L=c(fit$loadings.pm[[1]][,1:showK]), 
           x=rep(1:80),
           factor=as.factor(rep(1:showK, 
                      each=80))) %>%
  ggplot()+geom_line(aes(x=x, y=L, col=factor))
}
```

## flashier on `cov(t(X))` with point exponential prior

### data `X2`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(X2)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
fit1$pve
plot.factors(fit1, 3)

# with backfit
fit1b <- flash(cov(t(X2)), verbose.lvl=0, 
              backfit=TRUE, prior.family = prior.point.exponential())
fit1b$pve
plot.factors(fit1b, 3)
```

### data `X2n_a`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(X2n_a)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
fit1$pve
plot.factors(fit1, 5)

# with backfit
# fit1b <- flash(cov(t(X2n_a)), verbose.lvl=0, 
#               backfit=TRUE, prior.family = prior.point.exponential())
# returns error: Standard errors must be positive (and nonzero).
```

### data `X2n_b`

```{r}
# wihtout backfit
fit1 <- flash(cov(t(X2n_b)), verbose.lvl=0, 
              backfit=FALSE, prior.family = prior.point.exponential())
fit1$pve
plot.factors(fit1, 5)

# with backfit
fit1b <- flash(cov(t(X2n_b)), verbose.lvl=0, 
              backfit=TRUE, prior.family = prior.point.exponential())
fit1b$pve
plot.factors(fit1b, 5)
```
